---
title: "Tutorial: Bulk RNA-seq Synthetic Data"
format: html
---

```{r}
# Libraries required for this
library(tidyverse)
# Set up data/ directory
dir.create("data", showWarnings = FALSE)
```

We're going to create a fictional case to explain RNA-seq.

Let's say we have some pathogenic microorganism _V. scarious_, and we're 
trying to understand how chemicals can kill it. 
We've done bulk RNA-seq, n=4, with two conditions, one WT, and one
with a chemical compound X, and we want to see how the organism responds.

For this simulated data, we'll say 10 genes respond to the chemical, because
the organism's metabolic pathways are disrupted, and it needs to turn on
alternative pathways.

- Some of those genes are highly expressed, whilst other aren't. 
    - 
- 5 go down, and 5 go up. 

The "true" concentrations, in terms of reads, are as follows:

| Gene   | WT_reads | CompoundX_reads | Regulation   |
|--------|----------|-----------------|--------------|
| aceA   | 500      | 1200            | Up           |
| pykF   | 800      | 1600            | Up           |
| gltA   | 400      | 410             | Up           |
| crs    | 3        | 5               | Up           |
| tpiA   | 3        | 1000            | Up           |
| icd    | 1500     | 700             | Down         |
| pgi    | 900      | 400             | Down         |
| fumA   | 600      | 200             | Down         |
| lpdA   | 300      | 100             | Down         |
| zwf    | 20       | 15              | Down         |

- Let's say we also have 90 genes that don't respond to the condition, with
various expression levels.

We'll add those as follows:

| Gene   | WT_reads | CompoundX_reads | Regulation   |
|--------|----------|-----------------|--------------|
| nrsX   | 1..1000  | idem            | None         |

Where `1..1000` is a random integer number between 1 and 1000, that is equal
betwen the two conditions.

Now, let's put this an R table

```{r}

# Table with DE gene reads
df_gene_data_resp <- data.frame(
    Gene = c("aceA", "pykF", "gltA", "crs", "tpiA", "icd", "pgi", "fumA", "lpdA", "zwf"),
    WT.rep1 = c(500, 800, 400, 3, 3, 1500, 900, 600, 300, 20),
    CompoundX.rep1 = c(1200, 1600, 410, 5, 1000, 700, 400, 200, 100, 15)
    # Regulation = c("Up", "Up", "Up", "Up", "Up", "Down", "Down", "Down", "Down", "Down")
)
# Table with 90 non-responding genes
expression_levels_nresp90 <- sample(1:1000, 90, replace = TRUE) 
df_gene_data_extra <- data.frame(
    Gene = paste0("nrs", 11:100), # nrs = non responsive
    WT.rep1 = expression_levels_nresp90,
    CompoundX.rep1 = expression_levels_nresp90
)

# merge the two
df_gene_data <- bind_rows(df_gene_data_resp, df_gene_data_extra)

# add a gene that makes the total reads equal
max_totalreads = max(colSums(df_gene_data[,-1]))
hk101_expression = 1000+max_totalreads-colSums(df_gene_data[,-1])
df_gene_data <- df_gene_data %>%
    add_row(Gene = "hk101", 
            WT.rep1 = hk101_expression[1], 
            CompoundX.rep1 = hk101_expression[2])

# Now expand each of the reads columns 4x (to emulate replicates)
df_gene_data_all <- df_gene_data %>%
    mutate(
        WT.rep2 = WT.rep1,
        WT.rep3 = WT.rep1,
        WT.rep4 = WT.rep1,
        CompoundX.rep2 = CompoundX.rep1,
        CompoundX.rep3 = CompoundX.rep1,
        CompoundX.rep4 = CompoundX.rep1
    ) 
# Re-arrange the columns, put all WT first then the CompoundX cols
df_gene_data_all <- df_gene_data_all %>%
    select(Gene, starts_with("WT"), starts_with("CompoundX"))


head(df_gene_data_all)   
colSums(df_gene_data_all[,-1])

```

Let's see how this would look as a heatmap:

```{r}

library(pheatmap)

# 
df_annotation <- data.frame(
    row.names = colnames(df_gene_data_all[,-1]),
    Condition = rep(c("WT", "CompoundX"), each = 4)
)

# Prepare the data for pheatmap
heatmap_data <- df_gene_data_all %>%
    column_to_rownames(var = "Gene") %>%
    as.matrix()
    
# Generate the heatmap
# note that we cannot scale, because variance = 0 for non-bg
pheatmap(heatmap_data,
            cluster_rows = FALSE, 
            cluster_cols = FALSE, 
            show_rownames = TRUE, 
            show_colnames = TRUE,
            scale="none",
            annotation_col = df_annotation
)
    
# Show the DE genes  
pheatmap(heatmap_data[1:10,],
            cluster_rows = FALSE, 
            cluster_cols = FALSE, 
            show_rownames = TRUE, 
            show_colnames = TRUE,
            scale="row",
            annotation_col = df_annotation
)            

```


Now convert the `df_gene_data_all` to a dataframe with negative binomial noise.

RNA-seq counts are typically modeled with a **negative binomial distribution**,
which has variance $\text{Var} = \mu + \mu^2 / \theta$, where $\mu$ is the mean
and $\theta$ (`size`) is the dispersion parameter. A larger $\theta$ means less
noise (approaching Poisson). We'll use $\theta = 10$, a realistic moderate level.

```{r}

set.seed(42) # for reproducibility

dispersion_size <- 100  # size parameter (theta); higher = less noise

# Apply NB noise to each count value
df_gene_data_noisy <- df_gene_data_all %>%
    mutate(across(-Gene, ~ ifelse(.x > 0, 
                                   rnbinom(n(), mu = .x, size = dispersion_size), 
                                   0)))

head(df_gene_data_noisy)

```

Now let's inspect this data:

```{r}
df_gene_data_noisy %>%
    column_to_rownames(var = "Gene") %>%
    as.matrix() %>%
    pheatmap(cluster_cols = FALSE, 
             annotation_col = df_annotation)
```

And export this data to a tsv file, as would be typical for a count table, 
and also export metadata.



```{r}

# Export count matrix (genes as rows, samples as columns, no row names column)
# DESeq2 expects integer counts with gene IDs as row names
df_counts <- df_gene_data_noisy %>%
    column_to_rownames(var = "Gene")

write.table(df_counts, 
            file = "data/count_table.tsv", 
            sep = "\t", 
            quote = FALSE, 
            col.names = NA)  # col.names=NA puts a blank in the top-left corner

# Export sample metadata (colData for DESeq2)
# Row names must match the column names of the count matrix
df_metadata <- data.frame(
    sample = colnames(df_counts),
    condition = rep(c("WT", "CompoundX"), each = 4)
)

write.table(df_metadata, 
            file = "data/sample_metadata.tsv", 
            sep = "\t", 
            quote = FALSE, 
            row.names = FALSE)

cat("Exported count_table.tsv and sample_metadata.tsv\n")

```