---
title: "Tutorial: DESeq2 Analysis on Synthetic Data"
format: html
output-dir: html
---

## Introduction

In this tutorial we perform a standard **DESeq2** differential expression
analysis on the synthetic bulk RNA-seq data generated in the previous
tutorial. We will:

1. Load the count table and sample metadata
2. Create a DESeq2 dataset and run the analysis
3. Inspect the data with a PCA plot
4. Examine the dispersion–mean relationship
5. Extract the results table of differentially expressed genes (DEGs)
6. Visualise the results with a volcano plot

```{r}
#| warning: false

library(tidyverse)
library(DESeq2)
library(pheatmap)
library(ggrepel)
```

## 1 — Load the data

We read in the count matrix and metadata that were exported in the
data-generation tutorial.

```{r}

# Count matrix — first column becomes row names
count_data <- read.table("data_frozen/count_table.tsv", 
                         sep = "\t", 
                         header = TRUE, 
                         row.names = 1, 
                         check.names = FALSE)

# Sample metadata
col_data <- read.table("data_frozen/sample_metadata.tsv", 
                       sep = "\t", 
                       header = TRUE)

# DESeq2 requires the row names of colData to match the column names of the counts
rownames(col_data) <- col_data$sample

# Make sure condition is a factor with WT as the reference level
col_data$condition <- factor(col_data$condition, levels = c("WT", "CompoundX"))

head(count_data)
col_data

```

## 2 — Create the DESeqDataSet and run DESeq2

`DESeqDataSetFromMatrix` ties together the count matrix and the metadata.
The **design formula** `~ condition` tells DESeq2 we want to test for
differences between WT and CompoundX.

```{r}

dds <- DESeqDataSetFromMatrix(countData = round(count_data),
                              colData   = col_data,
                              design    = ~ condition)

# Run the full DESeq2 pipeline:
#   estimation of size factors, dispersion, and model fitting
dds <- DESeq(dds)

```

## 3 — PCA plot

A Principal Component Analysis gives us a quick overview of how the
samples cluster. We expect the WT and CompoundX samples to separate
along one of the principal components.

We use the **variance-stabilising transformation** (`vst`) so that the
PCA is not dominated by a few highly-expressed genes.

```{r}

# Normally, you'd use (for many genes)
# vsd <- vst(dds, blind = FALSE)
# Now, we'll use:
vsd <- varianceStabilizingTransformation(dds, blind = FALSE)

plotPCA(vsd, intgroup = "condition") +
    theme_minimal() +
    ggtitle("PCA — WT vs CompoundX")

```

## 4 — Dispersion estimates

DESeq2 models each gene's dispersion (biological variability) as a
function of its mean expression. The plot below shows:

- **black dots**: gene-wise dispersion estimates
- **red line**: the fitted trend
- **blue dots**: the final (shrunken) estimates used for testing

Genes with low counts are shrunk more towards the trend, borrowing
strength across genes.

```{r}

plotDispEsts(dds, main = "Dispersion estimates")

```

## 5 — Results table

We extract the results for the contrast **CompoundX vs WT**. Key columns:

| Column           | Meaning                                      |
|------------------|----------------------------------------------|
| `log2FoldChange` | log₂ fold-change (CompoundX / WT)            |
| `pvalue`         | raw p-value from the Wald test               |
| `padj`           | Benjamini–Hochberg adjusted p-value           |

```{r}

res <- results(dds, contrast = c("condition", "CompoundX", "WT"))

# Order by adjusted p-value
res_ordered <- res[order(res$padj), ]

# Show the top DEGs
head(as.data.frame(res_ordered), 20)

# How many genes pass padj < 0.05?
summary(res, alpha = 0.05)

```

## 6 — Volcano plot

A volcano plot shows **statistical significance** (−log₁₀ adjusted
p-value) on the y-axis against **effect size** (log₂ fold-change) on the
x-axis. Genes that are both statistically significant and have a large
fold-change appear in the top-left and top-right corners.

```{r}

# Convert results to a data frame for ggplot
df_volcano <- as.data.frame(res) %>%
    rownames_to_column("Gene") %>%
    mutate(
        significant = case_when(
            is.na(padj)            ~ "NS",
            padj < 0.05 & log2FoldChange >  1 ~ "Up",
            padj < 0.05 & log2FoldChange < -1 ~ "Down",
            TRUE                               ~ "NS"
        )
    )

# Top 5 up-regulated and top 5 down-regulated genes (by padj)
df_volcano_top <- df_volcano %>%
    filter(significant %in% c("Up", "Down")) %>%
    group_by(significant) %>%
    slice_min(padj, n = 5) %>%
    ungroup()

ggplot(df_volcano, aes(x = log2FoldChange, y = -log10(pvalue), colour = significant)) +
    geom_point(alpha = 0.7, size = 1.5) +
    scale_colour_manual(values = c("Up" = "firebrick", "Down" = "steelblue", "NS" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
    geom_vline(xintercept = c(-1, 1),     linetype = "dashed", colour = "grey40") +
    geom_text_repel(data = df_volcano_top, aes(label = Gene),
                    size = 3, max.overlaps = 20, show.legend = FALSE) +
    theme_minimal() +
    labs(title = "Volcano plot — CompoundX vs WT",
         x = expression(log[2]~fold~change),
         y = expression(-log[10]~p~value),
         colour = "Direction")

```

## Questions

We used generated data for this tutorial. These were the details we put in:

| Gene   | WT_reads | CompoundX_reads | Regulation   |
|--------|----------|-----------------|--------------|
| aceA   | 500      | 1200            | Up           |
| pykF   | 800      | 1600            | Up           |
| gltA   | 400      | 410             | Up           |
| crs    | 3        | 5               | Up           |
| tpiA   | 3        | 1000            | Up           |
| icd    | 1500     | 700             | Down         |
| pgi    | 900      | 400             | Down         |
| fumA   | 600      | 200             | Down         |
| lpdA   | 300      | 100             | Down         |
| zwf    | 20       | 15              | Down         |
| nrsX   | 1..1000* | idem            | None         |

_*)  1..1000 is a random number between 1 and 1000, equal for both conditions._

### Now answer the following questions

- Look at the results table, `res_ordered`:
    - When we look at the p-values column, ie `pvalue`, we see many genes with
    significant p-values. How come many of the "nrs" genes are also considered
    significant when we look at `pvalue`? 
        - What is p-value correction?
        - Does this succeed completely in its aim?
    - Why are not all genes that we know are DE not included in this list? (E.g. our 
    key transcription factor crs.)
    - What is our false discovery rate? And what is our power?
- Look at the Volcano plot:
    - Can you link each of the genes in the table with true values above
    to the Volcano plot? 
- Select all significant genes, for example `padj<0.05` from `res_ordered`. 
Can you explain this?
- Select a gene, and make a plot of its expression values, perhaps look into some
genes that show behavior you expected and did not expect (if any).



