---
title: "DESeq2 Deep Dive"
format: 
  html:
    toc: true
    toc-depth: 1      # how many heading levels to include
    toc-title: "Contents"
    toc-location: body # left, right, or body
editor: visual
knitr:
  opts_chunk:
    warning: false
---


# Setting up

## Acknowledgements

Aside from the [DESeq2 paper itself](http://doi.org/10.1186/s13059-014-0550-8) (and the earlier [DESeq paper](https://www.nature.com/articles/npre.2010.4282.1)), additional [explanation](https://hbctraining.github.io/Intro-to-DGE/schedule/links-to-lessons.html) from [Mistry et al.](https://doi.org/10.5281/zenodo.4783480) from the Harvard Chan Bioinformatics Core was very helpfull in compiling this material. In addition, I used Claude Opus 4.6's understanding of DESeq2, which I fact-checked with aforementioned resources and in the original [DESeq2 source code](https://github.com/thelovelab/DESeq2) (15f2ec9).

```{=html}
<!--
Meeta Mistry, Mary Piper, Jihe Liu, & Radhika Khetani. (2021, May 24). hbctraining/DGE_workshop_salmon_online: Differential Gene Expression Workshop Lessons from HCBC (first release). Zenodo. https://doi.org/10.5281/zenodo.4783481. RRID:SCR_025373.
-->
```

## Libraries

Import libraries that we'll need.

```{r}
#| warning: false

library(tidyverse)
library(DESeq2)
library(pheatmap)
library(ggrepel)
library(scales)
library(patchwork)

# Set up output directory
if (!dir.exists("plots/tut-3")) {
    dir.create("plots/tut-3", recursive = TRUE, showWarnings = FALSE)
}

```

Define some functions and settings to make our plots look fancy later.

```{r}

# Bars should be black per default
update_geom_defaults("bar", list(fill = "black"))  # sets it globally for session

# Custom adjust font function (to be used with ggplot2)
custom_adjust_fonts <- function(size = 15, family = "Arial") {
  theme(
    text             = element_text(size = size, family = family),
    axis.text        = element_text(size = size, family = family),
    axis.title       = element_text(size = size, family = family),
    plot.title       = element_text(size = size, family = family),
    plot.subtitle    = element_text(size = size, family = family),
    legend.text      = element_text(size = size, family = family),
    legend.title     = element_text(size = size, family = family),
    strip.text       = element_text(size = size, family = family)
  )
}

# Custom function to set bar graph settings I prefer (to be used with ggplot2)
custom_theme_bar <- function() {
  list(
    theme_classic(),
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust=.5),    
      plot.title = element_text(hjust = .5)
    ),
    scale_y_continuous(
      labels = scales::label_comma(), 
      expand = expansion(mult = c(0, 0.05)),
      limits = c(0, NA)
      ),
    custom_adjust_fonts()
  )
}

```

## Load data

Let's first load two datasets: - The synthetic one that was generated in tutorial 1. - Example dataset [GLDS38](https://osdr.nasa.gov/bio/repo/data/studies/OSD-38), for which pre-processed count data is available in the `data-real/` directory.

```{r}
# Import the table
count_data_synth <- read.table("data_frozen/count_table.tsv", 
                         sep = "\t", 
                         header = TRUE, 
                         row.names = 1, 
                         check.names = FALSE)
# Also import real data (spaceflight)
# (Downloaded from https://github.com/mishapaauw/RNA-seq-workshop/tree/main/data
# where Misha created a downsampled version of the original data.)
count_data_real <- read.csv("/Users/m.wehrens/Data_UVA/example-datasets/2026_RNAseq-Misha/GLDS38_raw_counts.csv",
                            row.names=1)
count_data_real_meta <- read.csv("/Users/m.wehrens/Data_UVA/example-datasets/2026_RNAseq-Misha/samples_to_condition.csv",
                          row.names=1)

TEST_RUN = TRUE
if (TEST_RUN) {
  # subset `count_data_real` to first 3000 rows
  count_data_real <- count_data_real[1:3000, ]
}
                          
# Now show 
```

And let's convert this to the tidy (aka 'long') format.

```{r}
# Convert synthetic data to long format
count_data_synth_long <- count_data_synth %>%
    rownames_to_column(var="gene") %>%
    pivot_longer(-gene, 
                names_to = "condition",
                values_to = "counts") 
    
# Convert real data to long format
count_data_real_long <- count_data_real %>%
    rownames_to_column(var="gene") %>%
    pivot_longer(-gene, 
                names_to = "sample",
                values_to = "counts") %>%

    # Additionally, set condition to FLT or GC, based on presence of _FLT_ or 
    # _GC_ string in sample name column
    mutate(condition = case_when(
        str_detect(sample, "_FLT_") ~ "FLT",
        str_detect(sample, "_GC_") ~ "GC",
        TRUE ~ sample
    )) |>
    # convert to factors with lvl order
    mutate(condition = factor(condition, levels=c("GC", "FLT"))) |>
    
    # And create abbreviated sample names
    mutate(sample_brief=str_remove(sample, "Atha_WT_Col_0_sl_")) |>
    mutate(sample_brief=str_remove(sample_brief, "_G.S.$")) 
```

# Overview of the DESeq2 appraoch

When running the DESeq2 analysis, we see the following output:

```         
estimating size factors
estimating dispersions
gene-wise dispersion estimates
mean-dispersion relationship
final dispersion estimates
fitting model and testing
```

<font color=red>

XXXXXXXXX

What happens during the DESeq2 analysis?

1.  **Normalization using size factors**
2.  **Dispersion estimate**
3.  **Estimate dispersion per gene**
4.  \*\*

</font>

# Normalization: size factors

### A simple approach

If sample 1 in total has 5000 reads, and sample 2 in total has 10000 reads
(a higher "sequencing depth"), and we consider a gene X, we expect that gene to 
have more reads in sample 2. In fact, a good guess would be that gene X has 2 times as many reads in sample 2.
This is of course not due to biology, but due to the fact that we sequenced
sample 2 more.

So, to more honestly compare samples, we need to *normalize* for the amount of reads in each samples.

When we have count table where rows correspond to genes, and columns to samples, we could simply divide each column by its total, and work with those values. Or, one step better, do that, and multiply with the mean of total sample counts, such that the values we end up with are in the same range as those with which we started.

### DESeq2's approach

DESeq2 takes a little more sophisticated appraoch. 
They make the argument that we're intersted in genes that show changes due to regulation. 

_"The total number of reads, $\sum_i k_{i,j}$, may seem to be a good measure of sequencing depth and hence a reasonable choice for $s_j$. Experience with real data, however, shows this not always to be the case, because a few highly and differentially expressed genes may have strong influence on the total read count, causing the ratio of total read counts not to be a good estimate for the ratio of expected counts."_
-- ([Anders & Huber, 2010)](https://doi.org/10.1186/gb-2010-11-10-r106))

When there are a few highly-abundant genes are upregulated, 
the concentrations of the other genes in the cells will go down, 
as the size of the total transcriptome remains similar. (And conversely,
if highly-abundant genes go down, other genes' concentrations will go up.)

DESeq2 thus tries to normalize both for sequencing depth, and this second order
effect.
It assumes that most gene don't change, and calculates 
the median change in counts over all genes as a 'size factor' that
reflects sequencing depth and the second order effect.
Changes in gene expression that remain left after normalization with
the size factor then should reflect change that we're intersted in.

Technically, DESeq2 calculates a pseudo-reference sample, with arithmetic averages over the samples for each gene, then normalizes each gene by the reference sample, and then calculates the median normalized expression per sample. This somewhat complex scheme is summarized below:

| Gene | sampleA | sampleB | Pseudo-reference sample | sampleA/ref | sampleB/ref |
|------------|------------|------------|----------------|------------|------------|
| EF2A | 1489 | 906 | sqrt(1489 \* 906) = 1161.5 | 1489/1161.5 = 1.28 | 906/1161.5 = 0.78 |
| ABCD1 | 22 | 13 | sqrt(22 \* 13) = 17.7 | 22/16.9 = 1.30 | 13/16.9 = 0.77 |
| ... | ... | ... | ... | ... | ... |
|  |  |  |  | **size factor sampleA:** median(1.28, 1.30, ..) = 1.30 | **size factor sampleB:** median(0.78, 0.77, ..) = 0.78 |

Let's go over this with some code.

## Total reads per sample

First, let's simply show the total reads per sample:

```{r}
# First determine total reads per sample
df_total_reads <- count_data_real_long |>
    group_by(sample) |>
    summarize(total_counts = sum(counts))
    
# Now plot in bar graph
p <- df_total_reads |>
  mutate(sample=str_remove(sample, "Atha_WT_Col_0_sl_")) |>
  ggplot(aes(x=sample, y=total_counts)) +
    geom_bar(stat="identity") +
    xlab('Sample')+ylab('Total counts')+
    custom_theme_bar() 
  
# save
ggsave("plots/tut-3/total_counts_per_sample.pdf", plot=p, width=10, height=16, units='cm', device=cairo_pdf)
# and show
p + custom_adjust_fonts(12)
```

## Estimate size factors manually

Now, estimate the size factors per sample, the same way DESeq2 does it.

```{r}

# Estimate the "size factors"
# Create reference sample; geometric mean
df_size_factors <- count_data_real_long |>  
  # Calculate geometric means of gene values
  group_by(gene) |>
  mutate(geom_mean = exp(mean(log(counts)))) |>
    # More explicit alternative; theoretically same, though less numerically stable
    # mutate(geom_mean = median(prod(counts)^(1/(length(counts))))) |>
  ungroup() |>
  # Filter out zero values (choice following DESeq2)
  filter(geom_mean > 0) |>
  # Now normalize the counts by geom mean
  mutate(counts_norm = counts/geom_mean) |>
  # And calculate median normalized counts (=size factor)
  group_by(sample) |>
  summarize(size_factor = median(counts_norm, na.rm=TRUE))

# now plot those
p2 <- df_size_factors |>
  mutate(sample=str_remove(sample, "Atha_WT_Col_0_sl_")) |>
  ggplot(aes(x=sample, y=size_factor)) +
  geom_bar(stat="identity") +
  xlab('Sample') + ylab('Size factor') +
  custom_theme_bar()
# save
ggsave("plots/tut-3/size_factors_per_sample.pdf", plot=p, width=10, height=16, units='cm', device=cairo_pdf)

# show side by side with totals
(p + custom_adjust_fonts(12) + ggtitle("Total read count")) + 
(p2 + custom_adjust_fonts(12) + ggtitle("Size factor"))

```

As you can see, relative changes in these quantities between samples are similar but not equal.

## Compare manual size factors with DESeq2

As a sanity check, let's see whether this matches DESeq2's calculation:

```{r}

# Let's see how well this matches DESeq2's assessment

# Set up a dds
dds <- DESeqDataSetFromMatrix(countData = count_data_real,
                              colData   = count_data_real_meta,
                              design    = ~ condition)
dds <- estimateSizeFactors(dds)
the_sizefactors <- sizeFactors(dds)
df_size_factors_deseq2 <- data.frame(
  size_factor = unname(the_sizefactors),
  sample = names(the_sizefactors))

# now comparison
df_size_factors_all <-  
  rbind( df_size_factors |> mutate(origin="manual"), 
        df_size_factors_deseq2 |> mutate(origin = "deseq2"))
        
# now plot
p <- df_size_factors_all |>
  mutate(sample = str_remove(sample, "Atha_WT_Col_0_sl_")) |>
  ggplot(aes(x = sample, y = size_factor, fill = origin)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_manual(values = c("manual" = "black", "deseq2" = "red")) + 
  xlab('Sample') + ylab('Size factor') +
  custom_theme_bar()
  
ggsave("plots/tut-3/size_factors_manual_vs_deseq2.pdf", plot = p, width = 10, height = 16, units = "cm", device=cairo_pdf)

p + custom_adjust_fonts(12)
```

The agreement between these bars shows that our manual calculation of the 
size factors exactly matches DESeq2's calculation.

## Manually normalize genes and see effect

Let's now apply normalization of the count values, and illustrate the effect with a plot for 10 random genes.

```{r}

# set seed for reproducibility
set.seed(123)

# add size factors to count table, and add normalized counts to table 
count_data_real_long_normalized <- count_data_real_long |>
  left_join(df_size_factors, by='sample') |>
  mutate(normalized_count = counts/size_factor)

# plot 10 random genes, with raw and normalized counts
p <- count_data_real_long_normalized |> 
  pivot_longer(cols=c(counts, normalized_count), names_to="type", values_to = "counts") |>
  filter(gene %in% sample(unique(gene), 9)) |>  
  ggplot(aes(x=sample_brief, y=counts, color=type, shape=condition)) +
    # geom_bar(stat="identity", position="dodge") +
    geom_point() +
    facet_wrap(~gene, scales = "free_y", ncol = 3) +
    theme_classic() +
    xlab('Condition') + ylab('Counts') +
    scale_color_manual(values = c("counts" = "black", "normalized_count" = "red")) + 
    # scale_color_manual(values = c("FLT" = "darkblue", "GC" = "brown")) +
    custom_theme_bar()

# and save
ggsave("plots/tut-3/normalized_counts_example_raw.pdf", plot=p, width=20, height=15, units='cm', device=cairo_pdf)

p + custom_adjust_fonts(12)

```

In the plot above, the red points better reflect the true transcript concentrations compared between the samples.

As a final point, note that the total counts for each sample are not equal after DESeq2's normalization.

```{r}
# Let's do a full run
dds_full <- DESeq(dds)

# We can now also get normalized count matrix from deseq2
normalized_counts <- counts(dds_full, normalized=TRUE)
# Which doesn't have all sample totals equal
colSums(normalized_counts)


p <- count_data_real_long_normalized |> 
  pivot_longer(cols=c(counts, normalized_count), names_to="type", values_to = "counts") |>
  group_by(sample_brief, type) |>
  summarize(total_counts = sum(counts)) |>
  ungroup() |> 
  ggplot(aes(x=sample_brief, y=total_counts, fill=type)) +
    geom_bar(stat="identity", position="dodge") +
    theme_classic() +
    xlab('Condition') + ylab('Total counts') +
    scale_fill_manual(values = c("counts" = "grey", "normalized_count" = "black")) + 
    custom_theme_bar() + 
    theme(legend.position = "none")

ggsave(
  "plots/tut-3/normalization_comparison.pdf", 
  plot=p, width=20, height=15, units='cm', device=cairo_pdf
  )

p + custom_adjust_fonts(12)

```

In the plot above, the grey colored bars indicate total counts per sample before 
normalization, the black bars indicate total counts after normalization.
As you can see, normalized counts (black) are (somewhat) more equal in height, 
but still different from each other.


# Calculate dispersions

## What's dispersion and why is it important

As mentioned, in the end, our aim is to know how changes in gene expression depend on the different conditions of our experiment.

Due to technical and biological factors, gene expression counts will vary, even within the same condition.
Dispersion (variability) refers to the extent by which observations of a quantity, in this case counts, vary.

In the example with gene A below, we can say with high likelihood that the gene is differentially expressed, whilst 
in the example with gene B, we can be less sure of that.

```{r}
#| code-fold: true

set.seed(123)

df_example_simpleDE <-
  data.frame(
    counts = c(
      20+rnorm(n = 3, mean = 10, sd = .25),
      20+rnorm(n = 3, mean = 15, sd = .25),
      20+rnorm(n = 3, mean = 10, sd = 15),
      20+rnorm(n = 3, mean = 15, sd = 15)
    ),
    gene = c(
      rep('gene A\n(low dispersion)', 6),
      rep('gene B\n(high dispersion)', 6)
    ),
    condition = c(
      rep('Control',3), 
      rep('Treatment',3), 
      rep('Control',3), 
      rep('Treatment',3)
    ))
    
p <- df_example_simpleDE |>
  ggplot(aes(x=condition, y=counts))+
  geom_point()+
  facet_grid(cols = vars(gene))+
  theme_classic() +
  ylab("Counts")+
  xlab("Condition")+
  ggtitle("n=3") +
  theme(plot.title = element_text(hjust = 1)) +
  custom_adjust_fonts()
  
# save 
ggsave("plots/tut-3/dispersion_example.pdf", plot=p, width=13, height=12, units='cm', device=cairo_pdf)

p + custom_adjust_fonts(12)

# Print fold-changes
df_FC <- df_example_simpleDE %>%
  group_by(gene, condition) %>%
  summarise(mean_count = mean(counts), .groups = "drop") %>%
  pivot_wider(names_from = condition, values_from = mean_count) %>%
  mutate(FC = Treatment/Control) 
  
print(paste0("Gene A has FC=", round(df_FC$FC[1],2),", ","Gene B has FC=", round(df_FC$FC[2],2)))

```

In these examples you can see by eye that there is a high variability
in gene B's expression, which is why we trust the fold change estimate much
less.

Trying to answer which genes are truely differentially expresssed, 
is our reason to use a tool like DESeq2, 
which will produce statistics to say something about this.

As you can see from the example above, dispersion (variability) plays a big role 
in determining whether we can trust the fold change estimate. 
Indeed, dispersion estimates play a central role in DESeq2's ability to
calculate stastistics that inform us about the extent and significance of 
differential gene expression.

## Estimating dispersion plays a large role in DESeq2

A load of code in DESeq2 is dedicated to estimate the dispersion (variability)
in each gene's expression that is not due to differences in experimental
conditions.
In broad strokes, this estimated dispersion for each gene is then used to:

- **Inform statistical testing.** 
At the end of the DESeq2 procedure, a p-value is calculated to determine statistical
significance of the differential expression. 
In the calculation of this p-value, an estimate of the dispersion is used.
- **Improve L2FC estimates.** 
DESeq2 will estimate the fold-change, which is based on the data, 
but variability estimates are used to refine the estimate.
- **Improve PCA accuracy.** 
Optionally, you can create a modified gene count matrix in which 
dispersion information is to normalize the dispersion. 
When you use a PCA for data visualization, basing it on the 
variance normalized ("variance-stabilized") count matrix, 
will result in a more accurate representation of the data by the PCA.

### The dispersion is not similar for all genes

Since the goal is to estimate dispersion that is not due to experimental conditions,
let's take a better look at the dispersion, and calculate some metrics that
quantify that dispersion.

```{r}

# First for the synthetic data
# now calculate mean, variance and sd
df_dispersion_synth <- count_data_synth_long %>%      
    group_by(gene) %>%
    summarise(mean = mean(counts),
              sd   = sd(counts),
              variance = var(counts)) %>%
    # also calculate "alpha", which we'll use later
    mutate(alpha = (variance-mean)/mean^2)
    
# Then for the real data
# now calculate mean, variance and sd    
df_dispersion_real <- count_data_real_long %>%      
    group_by(gene, condition) %>%
    summarise(mean = mean(counts),
              sd   = sd(counts),
              variance = var(counts)) %>%
    # also calculate "alpha", which we'll use later
    mutate(alpha = (variance-mean)/mean^2)
```

When we look at the values that we now have, we'll see that there is a wide
range of dispersion values per gene:

```{r}
# Histogram of dispersion values for all genes

p <- df_dispersion_real |>
    ggplot(aes(x=sd)) +
    geom_histogram(bins=100) +
    theme_classic() +
    scale_x_log10(labels = scales::label_comma()) +
    xlab("Standard deviation for a gene") +
    ylab("Times observed") +
    ggtitle("Spaceflight data") +
    custom_theme_bar()
    
ggsave("plots/tut-3/dispersion_histogram.pdf", plot=p, width=10, height=10, units='cm', device=cairo_pdf)

p + custom_adjust_fonts(12)

```

If we plot the standard deviation versus the mean, we'll see that there's a clear
dependency of the sd on the mean.

```{r}
# sd versus mean relationship

p <- df_dispersion_real |>
    ggplot(aes(x=mean, y=sd)) +
      # trendline
    geom_smooth(method='lm', formula = y ~ x, color='red') +
    # points
    geom_point() +
    # R = ..
    annotate("text", x = 0.1, y = max(df_dispersion_real$sd)*.95, 
             label = paste0("R = ", round(cor(df_dispersion_real$mean, df_dispersion_real$sd), 2)),
             hjust = 0, vjust = 1) +
    theme_classic() +
    scale_x_log10() + scale_y_log10() +
    xlab("Mean expression (log scale)") +
    ylab("Standard deviation (log scale)") +
    ggtitle("Spaceflight data") 
    
ggsave("plots/tut-3/dispersion_sd_vs_mean.pdf", plot=p, width=10, height=10, units='cm', device=cairo_pdf)

p + custom_adjust_fonts(12)

```

What DESeq2 will try to do is quantify this trend, 
and use that to estimate each gene's dispersion due to "randomness"
(which is then later used to calculated p-values, et cetera, as mentioned above).

Specifically, DESeq2 assumes gene count values show statistics that are 
consistent with a negative binominal distribution.
For such a distribution, the variance and mean are related by the following
relationship:

$$
\text{variance} = \mu + \alpha \cdot \mu^2
$$ 

With $\alpha$ a free parameter that determines how the relationship looks precisely.
Let's see how consistent the data is with this assumption:

```{r}
# Fit alpha

# The function describing variance = mu + alpha * mu^2
fn_NBvariance <- function(mu, alpha) {
    return(mu + alpha * mu^2)
}

# Let's first fit the dispersion coefficient
# Fit alpha in: variance = mu + alpha * mu^2
# Use nls on log-transformed values for a better fit across the wide range
df_fit <- df_dispersion_real %>% filter(mean > 0, variance > 0)
fit <- nls(log(variance) ~ log(mean + alpha * mean^2),
           data = df_fit,
           start = list(alpha = 0.01),
           control = nls.control(maxiter = 200))
MY_ALPHA_real <- coef(fit)[["alpha"]]
cat("Fitted alpha:", MY_ALPHA_real, "\n")

# Now plot
p <- ggplot(df_dispersion_real, aes(x=mean, y=variance)) +
    geom_point(color='black', alpha=.2) + 
    geom_density2d(color='black', bins=5) + 
    # geom_abline(slope=1, intercept=0, color='red', size=1) +    
    stat_function(fun=fn_NBvariance, args=list(alpha=MY_ALPHA_real), color='orange', size=1) +
    stat_function(fun=fn_NBvariance, args=list(alpha=MY_ALPHA_real), color='black', size=.25) +    
    scale_x_log10() + scale_y_log10() + 
    theme_classic() +
    annotate("text", x = 0.1, y = max(df_dispersion_real$variance)*.95, 
             label = paste0("variance = mean + alpha * mean^2\nalpha =",round(MY_ALPHA_real,2)),
             hjust = 0, vjust = 1) +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)') +
    ggtitle('Spaceflight data')

# export to directory plots/tut-3/
ggsave("plots/tut-3/dispersion_variance_mean_real_data.pdf", plot=p, width=10, height=10, units='cm')

p + custom_adjust_fonts(12)
```


This seems to work rather well for our data. Using this information, 
we can now estimate the variability of any gene, given its mean:

$$
\text{estimated variance}_\text{gene} = \mu_\text{gene} + \alpha_\text{global fit} \cdot \mu_\text{gene}^2
$$

This is great! If we now have a gene whos average expression is different between
two conditions, we can use this information to estimate how likely it is that
that difference was due to chance.

Let's illustrate this with a randomly selected gene (that only has few counts).

```{r}

set.seed(1234)

example_gene <- sample(
  (count_data_real_long %>% filter(counts > 10 & counts < 15))$gene,
  1
)

mean_GC <- 
  mean(
    (
      count_data_real_long %>% 
        filter(gene == example_gene & condition == 'GC')
    )$counts
  )

p <- count_data_real_long %>% 
  filter(gene == example_gene) %>%
  ggplot() +
    geom_jitter(aes(x = counts, y = 0, color=condition), height=.006, width=0, size=5) +
    scale_color_manual(values = c("GC" = "orange", "FLT" = "darkblue"))  +
    ylim(c(-.02, .1)) +
    theme_classic() +
    # add a gaussian with mu = mean_GC, and sd = sqrt(estimated variance for this gene)
    # geom_function(
    #   fun = function(x) dnorm(x, mean = mean_GC, sd = sqrt(fn_NBvariance(mean_GC, MY_ALPHA_real))),
    #   color = "orange",
    #   size = 1
    # ) +
    # 
    # .. but the model uses a neg binominal
    # (use geom_point at integer x values, since dnbinom is a discrete PMF)
    geom_line(
      data = data.frame(x = 0:60) |> mutate(y = dnbinom(x, mu = mean_GC, size = 1/MY_ALPHA_real)),
      aes(x = x, y = y),
      color = "orange",
      size = 1
    ) +
    xlim(c(0,60)) +
    ylab("Probability for negative binominal distribution") +
    xlab("Counts") +
    custom_adjust_fonts(15) + 
    ggtitle(paste0('Counts for ', example_gene))

# Save 
ggsave("plots/tut-3/statistics-gene-example.pdf", plot=p, width=10, height=10, units='cm', device=cairo_pdf)

# And show
p + custom_adjust_fonts(12)

```

Here, the x-axis shows the counts of the selected gene, and points indicate
specific values that were recorded (for points, differences in the y direction  are 
only for visualization purposes). 
The line indicates the negative binominal distribution with mean equal to the 
mean of the gene's expression in the GC condition, and dispersion as estimated by our fit.
To determine whether there is statistical evidence for differential expression,
we want to know the probability that the count values of the FLT conditions
occured only by chance. That is, given the yellow distribution (H0), what is 
the chance of finding these FLT values.

The probability getting a single value that's higher than a value x, e.g. $x = 21$,
is the integrated probability of the negative binominal.

```{r}
#| code-fold: true

# Calculate the probability that x > 21, assuming neg bin distribution
prob_x21 <- 1 - pnbinom(21, mu = mean_GC, size = 1/MY_ALPHA_real)

# And plot
p <- count_data_real_long %>% 
  filter(gene == example_gene) %>%
  ggplot() +    
    ylim(c(-.02, .1)) +
    theme_classic() +
    geom_area(
      data = data.frame(x = 21:60) |> mutate(y = dnbinom(x, mu = mean_GC, size = 1/MY_ALPHA_real)),
      aes(x = x, y = y),
      color = "orange",
      fill = "lightgrey",
      size = 1
    ) +
    geom_line(
      data = data.frame(x = 0:60) |> mutate(y = dnbinom(x, mu = mean_GC, size = 1/MY_ALPHA_real)),
      aes(x = x, y = y),
      color = "orange",
      size = 1
    ) +
    geom_point(aes(x = 21, y = 0, color=condition), height=.006, width=0, size=5, color="darkblue") +
    annotate("text", x = 22, y = max(dnbinom(0:60, mu = mean_GC, size = 1/MY_ALPHA_real))*0.1, 
             label = paste0("P(x > 21) = ", round(prob_x21, 2)),
             hjust = 0, vjust = 1) +
    xlim(c(0,60)) +
    ylab("Probability for negative binominal distribution") +
    xlab("Counts") +
    custom_adjust_fonts(15) + 
    ggtitle(paste0('Counts for ', example_gene))

# Save 
ggsave("plots/tut-3/statistics-gene-example-chance.pdf", plot=p, width=10, height=10, units='cm', device=cairo_pdf)

# And show
p + custom_adjust_fonts(12)

```

For AT1G14240 to have >21 counts, that probability is 0.84, so very likely
under H0.

To calculate the likelihood that the mean of multiple FLT count observations are higher (or lower)
than the GC mean (reference condition), 
the calculation is a bit more complicated.

This is however a common problem, and 
we can use a Wald or t-test here.

These are very related, they both have the same test statistic:

$$
\frac{\mu_{GC} - \text{mean}_{FLT}}{\sqrt{\text{var}}/\sqrt{N}}
$$

($\mu$ indicating our H0 population mean.)

This test statistic can be assumed to be normally distributed due to the central limit theorem.
Or, more specifically, in the t-test this test-statistic is assumed to follow the t-distribution,
which gets broader for small sample sizes, but converges to the normal distribution for large sample sizes.
In the Wald test, this test statistic is simply assumed to be 
normally distributed (in other words, the t-test is a refinement of the Wald test).

For small sample sizes, in principle, the Wald test is not so suitable. 
However, since DESeq2 doesn't base the estimated variability and 
expression difference (reflected by the fold change) on information only
from a single gene, it has more reliable variance estimates, and can use the Wald test, 
and get more statistical power.

Which we can calculate to be:

```{r}
#| code-fold: true

df_counts_gene_example <- count_data_real_long %>% filter(gene == example_gene)

# Calculate the test statistic
mean_FLT <- mean((df_counts_gene_example %>% filter(gene == example_gene & condition == 'FLT'))$counts)
mean_GC  <- mean((df_counts_gene_example %>% filter(gene == example_gene & condition == 'GC'))$counts)
var_both <- fn_NBvariance(mean_GC, MY_ALPHA_real)

test_statistic <- (mean_GC-mean_FLT) / (sqrt(var_both)/sqrt(dim(df_counts_gene_example)[1]/2))
print(paste0('The test statistic t is ', round(test_statistic, 2)))

# And we can calculate the probabality as follows
MAX_VAL <- 5
p <- ggplot() +
  geom_function(
    # gaussian with mean 0 and sd 1, which is the distribution of the t statistic under H0
    fun = function(x) dnorm(x, mean = 0, sd = 1),
    color = "orange",
    size = 1) +
  stat_function(
    # fun = function(x) dt(x, df=2),
    fun = function(x) dnorm(x, mean = 0, sd = 1),
    geom = "area",
    xlim = c(-1*MAX_VAL, -1 * test_statistic),
    # color = "orange",
    fill = 'grey',
    size = 1
  ) +
  stat_function(
    # fun = function(x) dt(x, df=2),
    fun = function(x) dnorm(x, mean = 0, sd = 1),
    geom = "area",
    xlim = c(test_statistic, MAX_VAL),
    # color = "orange",
    fill = 'grey',
    size = 1
  ) +
  geom_vline(xintercept =test_statistic, color="darkblue", size=1) +  
  xlim(c(-1*MAX_VAL,MAX_VAL))+
  theme_classic() +
  custom_adjust_fonts(15) +
  xlab("Test-statistic") +
  ylab("Probability")
  
# Save
ggsave("plots/tut-3/statistics-tstatistic.pdf", plot=p, width=10, height=10, units='cm', device=cairo_pdf)  

# Show
p + custom_adjust_fonts(12)

# Calculate the probability
p_H0 <- 2 * (1 - pt(abs(test_statistic), df=2))
print(paste0("Our two-tailed p-value is ", round(p_H0, 3)))
```

So, this shows that our p-value is 0.451, and we have to accept H0, 
and conclude that there is no statistical evidence for differential expression of this gene.

XXXXXXXXX

I WAS EDITING HERE

XXXXXXXXX


Since log space ..

.. what DESeq2 actually does (..) test statistic = beta/SE(beta)

It turns out that, if we calculate the log fold change




XXXXXXX I WAS EDITING HERE XXXXXXXX


DESeq2 tries to quantify this trend, combine it with information on the observed 
dispersion of a gene, and then produces an updated dispersion estimate for 
a each gene separately.

As mentioned, these dispersion estimates are then used to 
calculate p-values for the likelihood of H0, and thus to determine which genes are likely differentially expressed.

::: {.callout-note}

## Technical: DESeq2 doesn't using fit the normalized counts

DESeq2 works with negative binomial distributions, which describe the
statistics of discrete numbers.

After normalization of the counts by the size factors, the counts are not
integers any more. 

Under the hood, DESeq2 keeps working with the discrete counts.

The model is as follows:

$$
K_{ij} \sim \text{NB}(\mu_{ij}, \alpha_i)
$$
$$
\log(\mu_{ij}) = \log(s_j) + \mathbf{x}_j \boldsymbol{\beta}_i
$$

Which can be written as: 
$$
K_{ij} \sim \text{NB}( s_j e^{\mathbf{x}_j \boldsymbol{\beta}_i} , \alpha_i)
$$

Here, $K_{ij}$ are the raw counts for gene $i$ in sample $j$ that are fitted ("$y$" from "$y = \beta_0 + \beta_1 * x$"),
with $s_j$ the size factor (and $x_j$ the condition, and $\beta_i$ the coefficients that we want to estimate).

$$
e^{\mathbf{x}_j \boldsymbol{\beta}_i}
$$ 

represents our quantity of interest,
the size-corrected mean counts that we can compare between samples.
(The factor $s_j$ scales that term back to the raw counts.)

Thus, the size factors are included in the fit.

Technically, this looks like:

```r
MASS::glm.nb(counts ~ condition + offset(log(size_factor)), data = df)
```

:::


# Estimate log fold changes


XXXXXXXXXXXXXXXXXXXXXXXXX

THE ESTIMATION OF DISPERSIONS DOES COME FIRST, AND CAN BE INCORPORATED 
EASILY IN THE MANUAL GLM!!
```{R}

###### CODE BY CLAUDE ######

# theta = 1/alpha (DESeq2's dispersion alpha, not to be confused with MASS's theta)

# my_alpha <- 0.5  # your dispersion estimate
# glm(
#   normalized_count ~ condition_coded,
#   data = df,
#   family = MASS::negative.binomial(theta = 1 / my_alpha)
# )
```

XXXXXXXXXXXXXXXXXXXXXXXXX

In the end, our aim is to know how changes in gene expression depend on the different conditions of our experiment.

To understand changes in gene expression, we could simply calculate the 
ratio of average expression in one condition, compared to the other.

```{r}

# A simple direct calculation of log2 fold changes, using averages
df_l2fc_manual <- count_data_real_long_normalized |>
    # first gather averages
    group_by(condition, gene) |>
    summarise(gene_average = mean(round(normalized_count)), .groups = "drop") |>    
    # now determine FC
    filter(gene_average>0) |>
    group_by(gene) |>
    pivot_wider(names_from = condition, values_from = gene_average) |>
    mutate(FC = FLT/GC, 
            L2FC = log2(FC)) |>
    arrange(desc(L2FC))

head(df_l2fc_manual)
```

And let's plot the counts of the top 3 genes to see whether this makes sense:

```{r}

# Plot the top 3
p <- count_data_real_long_normalized |>
  filter(gene %in% df_l2fc_manual$gene[1:3]) |>
  mutate(gene = factor(gene, levels=df_l2fc_manual$gene[1:3])) |>
  ggplot(aes(x=condition, y=counts, color=condition)) +
    geom_point() +
    facet_wrap(~gene, scales = "free_y") +
    theme_classic() +
    xlab('Condition') + ylab('Average normalized count') +
    scale_color_manual(values = c("GC" = "brown", "FLT" = "darkblue")) + 
    custom_theme_bar() + 
    theme(legend.position = "none")    
    
# Save
ggsave("plots/tut-3/manualfitting_top3_genes.pdf", plot=p, width=20, height=8, units='cm', device=cairo_pdf)    
    
# Show    
p + custom_adjust_fonts(12)    
    
```

However, to be more precise and versatile (allowing for multiple conditions, e.g.), 
DESeq2 fits the log2 fold changes using a "GLM" (generalized linear model), 
which is an extension of a linear model that allows
for non-normal distributions (DESeq2 assumes a negative binominal model for gene counts).

For example, if we have the condition $x$ corresponding to either ground
control ($x=0$) or space flight ($x=1$), we can fit gene expression $q_i$ for 
any gene $i$ with

$$
q_i = \beta_{0,i} + \beta_{1,i} x
$$

The idea here is to fit $\beta$ such that it reflects the effect
the condition $x$ has on the final expression of gene $i$.

We can perform the GLM fitting procedure as follows:

```{r}

# Let's manually convert ("dummy-code") the condition to 0 (GC) or 1 (FLT)
count_data_real_long_normalized <- count_data_real_long_normalized |>
  mutate(condition_coded = ifelse(condition == "GC", 0, 1))

# We use split-apply to fit each gene separately
model <- count_data_real_long_normalized %>%
  split(.$gene) %>%
  lapply(
    function(df) {
      # We use try-catch to deal with edge-cases where there is no
      # convergence or other issues, e.g. due to low counts or low variability
      tryCatch(
        expr = 
          # The actual fitting (negative binominal model)
          MASS::glm.nb(
            # The "model", describing q = beta0 + beta1 * condition_coded
            formula = normalized_count ~ condition_coded,
            # Gene data
            data = df
          ),
        error = 
          # Return NULL in case of error
          function(e) NULL
      )
    })

# Now let's summarize the result by collecting coefficients into dataframe
# Note: this code could be more generalized, but now it's more explicit.
df_coefficients <- data.frame(
  gene = names(model),
  # go over genes using sapply
  coefficient = sapply(model, function(x) {
    # obtain coefficient for this gene, converting to log2 instead of log
    if (is.null(x)) NA_real_ else coef(x)[["condition_coded"]] 
  }),
  baselvl    = sapply(model, function(x) {
    # obtain offset (intercept) for this gene 
    if (is.null(x)) NA_real_ else coef(x)[["(Intercept)"]] 
  })
) |>
  mutate(l2.coefficient  = coefficient/log(2)) |>
  arrange(desc(l2.coefficient))

```

::: {.callout-note}

## Technical notes

`MASS::glm.nb()` uses a "link" function to couple the output
of the linear model to the prediction, and per default it takes 
the log function, such that what we're fitting will be
$log(y) = \beta_0 + \beta_1 * x_{\text{condition}}$ (f(z) = log(z) being the 
link function here).
In addition, `MASS::glm.nb()` (or `lm()` in general) will dummy-code 
the condition column, such that GC will be coded as 0, and
FLT will be coded as 1 (alphabetical order).
Both these default behaviors also apply in DESeq2.

:::

Since 
$$
\text{ln}(y_\text{FLC} / y_\text{GC}) = 
ln(e^{(\beta_0 + \beta_1 * x_\text{FLC})} / e^{(\beta_0 + \beta_1 * x_\text{GC})}) =
\beta_1
$$
with $y$ begin gene expression, the coefficient $\beta_1$ directly reflects the log fold change between conditions (and $\beta_1 / log(2)$ the L2FC).
Likewise, the intercept $\beta_0$ reflects the baseline expression (log scale) in the reference condition (GC).

Thus, we can now inspect the coefficients that we gathered, as they directly 
represent fold-change. Let's plot two genes with a big differential expression,
and the fits by the GLM.

```{r}

# Top 10 genes (l2.coefficient = beta_1 / log(2) = L2FC)
head(df_coefficients, n=10)
  
# Now calculate the predictions
df_predictions <- df_coefficients %>%
  # Add coded condition
  mutate(condition.val1 = 0, condition.val2 = 1) %>%
  pivot_longer(cols=c(condition.val1, condition.val2), names_to = "condition", values_to = "condition_coded") |>
  # Add condition
  mutate(condition = ifelse(condition_coded == "0", "GC", "FLT")) %>%
  # Add predictions
  mutate(prediction = exp(baselvl + coefficient * condition_coded))
  
# Gather plotting data (top 2 genes, add some plotting-related stuff)
df_plot_counts <- count_data_real_long_normalized |>
  filter(gene %in% df_coefficients$gene[1:2]) |>
  mutate(
    # set lvl order
    condition = factor(condition, levels=c("GC", "FLT")),
    # add condition_coded
    condition_coded = ifelse(condition == "GC", 0, 1)
  )
df_plot_pred <- df_predictions |>
  filter(gene %in% df_coefficients$gene[1:2]) |>
  mutate(condition = factor(condition, levels=c("GC", "FLT"))) |>
    # add label
  mutate(label = paste0(
    "y = exp(", round(baselvl, 2), " + ", round(coefficient, 2), "x)"
  ))
  
# .. and plot
p <- ggplot() +
  # Plot predictions
  geom_line(
    data = df_plot_pred, 
    aes(x=condition_coded, y=prediction+.1, group=gene), 
    stat='identity')+
  geom_label(
    data = df_plot_pred,
    aes(x = 0, y = Inf, label = label),
    hjust = 0.1, vjust = 1.3  # nudge inward from top-right corner
  ) +
  # plot the data
  geom_jitter(
    data = df_plot_counts,
    aes(x=condition_coded, y=normalized_count+.1, color=condition),
    width = 0.2, height = 0) +
  # grid
  facet_grid(cols = vars(gene)) +
  # cosmetics
  # set xtick labels to display 0 and 1 only
  scale_x_continuous(breaks = c(0, 1)) +
  scale_y_log10() +
  theme_classic() +
  custom_adjust_fonts(15) +
  ylab("Count") +
  theme(legend.position="none") +
  scale_color_manual(values = c("GC" = "brown", "FLT" = "darkblue")) 

# Save
ggsave("plots/tut-3/manual-GLM_top2_genes.pdf", plot=p, width=13, height=7.5, units='cm', device=cairo_pdf)

# And show
p + custom_adjust_fonts(12)

```

Finally, let's do a sanity check.
Are our L2FC values from the fit similar to those reported by DESeq2?

```{r}

# Let's also retrieve the L2FC from DESeq2
df_results_combined <- 
  as.data.frame(
    # obtain DESeq2 results, FLT relative to GC
    results(dds_full, contrast=c("condition", "space_flight", "ground_control"))
    ) |>
  rownames_to_column(var="gene") |>
  left_join(df_coefficients, by='gene')
  
# And compare
p1 <- df_results_combined  |>
  ggplot(aes(x=log2FoldChange, y=l2.coefficient)) +
    geom_point() +
    theme_classic() +
    xlab("DESeq2 log2 fold change") +
    ylab("Coefficient from manual fit") +
    geom_abline(slope=1, intercept=0, color='red', label="x=y") +
    ggtitle("All data") +
    custom_adjust_fonts()
    
# Set bounds to main part of data    
coeff_low  = -5 # quantile(df_results_combined$coefficient, .05, na.rm=TRUE)
coeff_high = 5  # quantile(df_results_combined$coefficient, .95, na.rm=TRUE)
p2 <- p1 + 
  xlim(c(coeff_low, coeff_high)) + 
  ylim(c(coeff_low, coeff_high)) +
  coord_fixed() + 
  ggtitle(paste0("L2FC range ", round(coeff_low,1), " to ", round(coeff_high,1)))
    
p <- p2 + p1    
    
# and save
ggsave("plots/tut-3/manualfitting_log2fc_comparison.pdf", plot=p, 
  width=20, height=10, units='cm', device=cairo_pdf)
  
p + custom_adjust_fonts(12)  
```

Yes! It appears so. There are some deviations, which are probably 
due to DESeq2's better handling of edge cases (e.g. low counts, low variability).

And lastly, as expected, this also compares well to the manual 
calculation of the L2FC, which can be done easily in this case, 
as we only have two conditions.


```{r}
    
# Add this to the comparison frame as well
df_results_combined <- df_results_combined |>
    left_join(df_l2fc_manual, by='gene')

# and compare our fitting to our average l2fc in a scatter plot
range_min = min(df_results_combined$L2FC, na.rm=TRUE)
range_max = max(df_results_combined$L2FC, na.rm=TRUE)
p <- df_results_combined  |>
  ggplot(aes(x=L2FC, y=l2.coefficient)) +
    geom_point() +
    theme_classic() +
    xlab("Manual log2 fold change (average)") +
    ylab("Coefficient from manual fit") +
    geom_abline(slope=1, intercept=0, color='red', label="x=y") +
    ggtitle("All data") +
    custom_adjust_fonts() +
    xlim(c(range_min, range_max)) +
    ylim(c(range_min, range_max))

p
```


::: {.callout-note}

## Shrunken L2FC estimates

"A common difficulty in the analysis of \[high-throughput sequencing\] data is 
the strong variance of LFC estimates for genes with low read count. 
(..) weakly expressed genes seem to show much stronger differences between the 
\[conditions\] than strongly expressed genes"
-- ([Love, Huber & Anders, 2014](http://doi.org/10.1186/s13059-014-0550-8))

To counter the effect of blown-up L2FC values in lowly-expressed genes,
DESeq2 offers a correction on top of the L2FC estimated by the GLM.
The function `lfcShrink()` applies 
a "Bayesian" statistics approach, 
meaning that they re-calculate the estimated L2FC, taking into account
the assumption (and corresponding probability distribution) that 
most genes will have an L2FC of zero, and substantial variability around that
zero.

More precisely, a prior is used that is centered at zero (assumption: most genes aren't DE).
Its width is learned from the data, such that genes with weak evidence (low counts → wide likelihood) 
get pulled more strongly toward zero whilst genes with strong evidence (high counts → narrow likelihood) 
are barely affected (this is called an empirical Bayes approach, as 
the prior is based on the data itself).

:::







# Calculating p-values (statistical testing)

To achieve this, simply put, DESeq2 will estimate the dispersion (variability) for each gene
that is attributed to "randomness", such that it can estimate whether 
observed changes lie outside the expected random variability, and are thus likely to be 
related to the different experimental conditions.

Or, in more technical terms, it will calculate p-values based on dispersion estimates, 
which reflect the likelihood that H0 (no difference between conditions),
is true. 
If that probability is low (typically $p<0.05$ or $<0.01$), 
we can reject H0, and say that there is likely a difference between conditions.

The method by which this is done is quite a bit more complicated, but that's the 
gist of it.






# Text to still re-structure






# Fitting gene expression to conditions

EXPLAIN HOW EXPRESSION-CONDITION MODEL IS USED

FIRST FOR ESTIMATING LFC
THEN FOR DISPERSION ESTIMATES

Throughout 


XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX
add mean-dispersion relationship
quantified in different ways


## DESeq2's handling of dispersion

Estimating dispersion (variabillity) play a big role in DESeq2's approach. Four of DESeq2's messages refer to this process.

```         
estimating dispersions
gene-wise dispersion estimates
mean-dispersion relationship
final dispersion estimates
```

The first output announces the start of the dispersion section[^Explanation Claude, and see methods.R (`estimateDispersions.DESeqDataSet <- function(...`), core.R in DESeq2 [source](https://github.com/thelovelab/DESeq2) 15f2ec9.], the 
three that follow refer to sub-steps of that procedure.

I don't pretend to understand the DESeq2 code in full, but we can emulate 
the process that DESeq2 follows to understand what's going on.







```{r}
plotDispEsts(dds_full)
```


```{r}
# Information about the dispersion of gene counts plays a big role in
# the DESeq2 analysis.
#
# Dispersion (~variability) can be quantified in multiple way.
# 
# Here, we calculate the sd and mean, and additionally "alpha".
# Alpha is (variance-mean)/mean^2).

# this script yields
# - df_dispersion_synth
# - df_dispersion_real
```



# Variance, mean, and dispersion

## Variance versus mean

```{r}
# DESeq2 has a dispersion estimate parameter, alpha, defined by
# var(K) = mu + alpha * mu^2
# Thus
# alpha = (var(K)-mu) / mu^2
              
# Now plot variance versus mean
#
# Very nice source of information:
# https://hbctraining.github.io/DGE_workshop/lessons/01_DGE_setup_and_overview.html#rna-seq-count-distribution
# The red line gives the poisson relationship of the mean and variance
# The blue line gives the negative binominal relationship between mean and variance
fn_NBvariance <- function(mu, alpha) {
    return(mu + alpha * mu^2)
}
MY_ALPHA = 1/100
ggplot(df_dispersion_synth, aes(x=mean, y=variance)) +
    geom_point() + 
    geom_abline(slope=1, intercept=0, color='red') +
    stat_function(fun=fn_NBvariance, args=list(alpha=MY_ALPHA), color='blue') +
    scale_x_log10() + scale_y_log10() + 
    theme_bw() +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)')
```

## Dispersion estimate plot

```{r}
# Now plot (var(K)-mu) / mu^2 versus mean
# This produces a similar plot to the dispresion estimate plot of DESeq2
ggplot(df_dispersion_synth, aes(x=mean, y=alpha)) +
    geom_point() +
    scale_x_log10() + scale_y_log10(limits=c(1e-8, 10))+
    xlab("mean") + ylab("dispersion estimate\nalpha = (var(K)-mu) / mu^2")

# Note furthermore that DESeq2 selects the top-variance genes for taking 
# along in the PCA. 
```

## Fit dispersion on real data

(..)

# Log2 fold change

```{r}
# Calculate means and log2fc per condition
count_data_real_long_l2fc <- count_data_real_long |>
    # first gather averages
    group_by(condition, gene) |>
    summarise(gene_average = mean(counts)) |>    
    # now determine FC
    filter(gene_average>0) |>
    group_by(gene) |>
    pivot_wider(names_from = condition, values_from = gene_average) |>
    mutate(FC = FLT/GC, 
            L2FC = log2(FC)) |>
    arrange(desc(L2FC))
```

```{r}
# now make a small plot for gene="AT2G41355"
MYGENE = "AT2G41355"
p <- count_data_real_long |> 
    filter(gene==MYGENE) |>
    ggplot(aes(x=condition, y=counts))+
        geom_jitter(height = 0, width = 0.1) +
        theme_classic() +
        ggtitle(MYGENE)
ggsave("plots/tut-3/gene_example_highl2fc.pdf", plot=p, width=7, height=7, units='cm')
```

# PCA fundamentals

## PCA on synthetic count data

```{r}
# PCA expects samples as rows, genes as columns, so we transpose.
# We also scale and center the data (z-scores), which is standard practice.

pca_result <- prcomp(t(count_data_synth), center = TRUE, scale. = TRUE)

# How much variance does each PC explain?
pca_var <- summary(pca_result)$importance
print(pca_var[, 1:4])
```

## Scree plot

```{r}
# Plot: variance explained per PC
df_var_explained <- data.frame(
    PC = paste0("PC", 1:ncol(pca_var)),
    variance_explained = pca_var["Proportion of Variance", ]
) |> mutate(PC = factor(PC, levels = PC))

p_scree <- ggplot(df_var_explained, aes(x = PC, y = variance_explained)) +
    geom_bar(stat = "identity") +
    theme_classic() +
    ylab("Proportion of variance explained") +
    ggtitle("Scree plot - synthetic data")
p_scree
```

## PCA scatter (PC1 vs PC2)

```{r}
df_scores <- as.data.frame(pca_result$x) |>
    mutate(sample = rownames(pca_result$x),
           condition = ifelse(str_detect(sample, "^WT"), "WT", "CompoundX"))

p_pca <- ggplot(df_scores, aes(x = PC1, y = PC2, color = condition)) +
    geom_point(size = 3) +
    theme_classic() +
    xlab(paste0("PC1 (", round(pca_var["Proportion of Variance", 1] * 100, 1), "%)")) +
    ylab(paste0("PC2 (", round(pca_var["Proportion of Variance", 2] * 100, 1), "%)")) +
    ggtitle("PCA - synthetic data")
p_pca
```

## Gene loadings

```{r}
# The rotation matrix contains the loadings (weights) per gene per PC.

df_loadings <- as.data.frame(pca_result$rotation) |>
    rownames_to_column("gene") |>
    mutate(PC1_abs = abs(PC1), PC2_abs = abs(PC2))

# Top 15 genes by PC1 loading
top_pc1 <- df_loadings |> arrange(desc(PC1_abs)) |> head(15)
print("Top genes contributing to PC1:")
print(top_pc1 |> select(gene, PC1))

# Top 15 genes by PC2 loading
top_pc2 <- df_loadings |> arrange(desc(PC2_abs)) |> head(15)
print("Top genes contributing to PC2:")
print(top_pc2 |> select(gene, PC2))

# Plot loadings for PC1
p_loadings <- top_pc1 |>
    mutate(gene = fct_reorder(gene, PC1)) |>
    ggplot(aes(x = gene, y = PC1)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_classic() +
    ggtitle("Top 15 gene loadings on PC1")
p_loadings
```

## tpiA loading across all PCs

```{r}
df_tpiA <- df_loadings |>
    filter(gene == "tpiA") |>
    select(gene, starts_with("PC"), -PC1_abs, -PC2_abs) |>
    pivot_longer(-gene, names_to = "PC", values_to = "loading") |>
    mutate(PC = factor(PC, levels = paste0("PC", 1:8)))

p_tpiA <- ggplot(df_tpiA, aes(x = PC, y = loading)) +
    geom_bar(stat = "identity") +
    theme_classic() +
    ggtitle("tpiA loading across all PCs")
p_tpiA
```

# PCA on normalized data

```{r}
# PCA on normalized (variance-stabilized) data
# This builds on the normalization from the size factors section,
# re-creating the DESeq2 object to then apply VST and run PCA.
```

## Variance vs mean (raw)

```{r}
# There tends to be a trend in the variance, compared to mean expression values.
p <- ggplot(df_dispersion_real, aes(x=mean, y=variance)) +
    geom_point(color='grey') + 
    scale_x_log10() + scale_y_log10() + 
    theme_classic() +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)') +
    ggtitle('Spaceflight data')
p
ggsave("plots/tut-3/variance_mean_real_data_raw.pdf", plot=p, width=10, height=10, units='cm')
```

## Z-score normalization

```{r}
# If we were to use z-scores
# (Often done when using PCAs)
count_data_real_long_zscore <- count_data_real_long |>
  group_by(gene) |>
  mutate(counts_z = counts/sd(counts)) |>
  ungroup()
# Plot our gene
MYGENE = "AT2G41355"
p <- count_data_real_long_zscore |> 
    filter(gene==MYGENE) |>
    ggplot(aes(x=condition, y=counts_z))+
        geom_jitter(height = 0, width = 0.1) +
        theme_classic() +
        ggtitle(MYGENE)
p
```

```{r}
# now calculate dispersions z
df_dispersion_real_zscore <- count_data_real_long_zscore |>
    group_by(gene) |>
    summarise(mean = mean(counts_z),
              sd   = sd(counts_z),
              variance = var(counts_z)) |>
    mutate(alpha = (variance-mean)/mean^2)
# now plot dispersions z
p <- ggplot(df_dispersion_real_zscore, aes(x=mean, y=variance)) +
    geom_point(color='grey') + 
    scale_x_log10() + scale_y_log10() + 
    theme_classic() +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)') +
    ggtitle('Spaceflight data, z-scores')
p
ggsave("plots/tut-3/variance_mean_real_data_zscore.pdf", plot=p, width=10, height=10, units='cm')
```

```{r}
# similarly, we'd see a trend if we'd take the log
df_disperson_log <- count_data_real_long_zscore |>
    mutate(      
      log_counts   = log(counts+.1),
      log_counts_z = log(counts_z+.1)) |>
    group_by(gene) |>
    summarise(
      mean_z = mean(log_counts_z),
      sd_z   = sd(log_counts_z),
      variance_z = var(log_counts_z),
      mean = mean(log_counts), 
      sd = sd(log_counts),
      variance = var(log_counts))
# plot again
p <- ggplot(df_disperson_log, aes(x=mean, y=variance)) +
    geom_point(color='grey') + 
    scale_x_log10() + scale_y_log10() + 
    theme_classic() +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)') +
    ggtitle('Spaceflight data')
p
```

## Variance-stabilizing transformation (VST)

```{r}
# So now let's check out how dispersions look using vst
vsd <- vst(dds)
vst_counts <- assay(vsd)
df_dispersion_ddsvst <- vst_counts |>
    as.data.frame() |>
    rownames_to_column("gene") |>
    pivot_longer(-gene, names_to = "sample", values_to = "counts") |>
    group_by(gene) |>
    summarise(mean = mean(counts),
              sd   = sd(counts),
              variance = var(counts)) |>
    mutate(alpha = (variance - mean) / mean^2)
# now plot
p <- ggplot(df_dispersion_ddsvst, aes(x=mean, y=variance)) +
    geom_point(color='grey') + 
    scale_x_log10() + scale_y_log10() + 
    theme_classic() +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)') +
    ggtitle('Spaceflight data, variance stabilized')
p
ggsave("plots/tut-3/variance_mean_real_data_vst.pdf", plot=p, width=10, height=10, units='cm')
```

## PCA on VST data

```{r}
# now a pca plot by deseq2
pca_res <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
p <- ggplot(pca_res, aes(x=PC1, y=PC2, shape=condition,
                      color=condition)) +
    geom_point(size=3) +
    theme_classic() +
    xlab(paste0("PC1 (", round(100*pca_res$percentVar[1], 1), "%)")) +
    ylab(paste0("PC2 (", round(100*pca_res$percentVar[2], 1), "%)")) +
    ggtitle("PCA on vst data") +
    scale_color_manual(values=c("space_flight"="darkblue", "ground_control"="brown")) +
    theme(legend.position = "none")
p
ggsave("plots/tut-3/pca_deseq2.pdf", plot=p, width=10, height=10, units='cm')
```

## Normalized count totals (recap)

```{r}
# Also in the overall DESeq2 analysis, dispersion-stablization is used
# (See also https://hbctraining.github.io/Intro-to-DGE/lessons/04b_DGE_DESeq2_analysis.html)

# So we have
# - Size factor correction
# - Dispersion correction

# Plot, remind ourselves that also the per-sample count normalization went well
normalized_counts_sampletotals <- colSums(normalized_counts)
p <- 
  data.frame(
    counts = unname(normalized_counts_sampletotals),
    sample = names(normalized_counts_sampletotals)
  ) |>
  mutate(sample = str_remove(sample, "Atha_WT_Col_0_sl_")) |>
  ggplot(aes(x = sample, y = counts)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("Sample") + ylab("Counts\n(normalized by size factors)")
ggsave("plots/tut-3/total_counts_per_sample_normalized.pdf", plot=p, width=7, height=10, units='cm')


```

# Statistical testing

```{r}
```