---
title: "Tutorial: DESeq2 Analysis on Synthetic Data"
format: html
---

## Introduction

In this tutorial we perform a standard **DESeq2** differential expression
analysis on the synthetic bulk RNA-seq data generated in the previous
tutorial. We will:

1. Load the count table and sample metadata
2. Create a DESeq2 dataset and run the analysis
3. Inspect the data with a PCA plot
4. Examine the dispersion–mean relationship
5. Extract the results table of differentially expressed genes (DEGs)
6. Visualise the results with a volcano plot

```{r}
#| warning: false

library(tidyverse)
library(DESeq2)
library(pheatmap)
library(ggrepel)
```

## 1 — Load the data

We read in the count matrix and metadata that were exported in the
data-generation tutorial.

```{r}

# Count matrix — first column becomes row names
count_data <- read.table("data_frozen/count_table.tsv", 
                         sep = "\t", 
                         header = TRUE, 
                         row.names = 1, 
                         check.names = FALSE)

# Sample metadata
col_data <- read.table("data_frozen/sample_metadata.tsv", 
                       sep = "\t", 
                       header = TRUE)

# DESeq2 requires the row names of colData to match the column names of the counts
rownames(col_data) <- col_data$sample

# Make sure condition is a factor with WT as the reference level
col_data$condition <- factor(col_data$condition, levels = c("WT", "CompoundX"))

head(count_data)
col_data

```

## 2 — Create the DESeqDataSet and run DESeq2

`DESeqDataSetFromMatrix` ties together the count matrix and the metadata.
The **design formula** `~ condition` tells DESeq2 we want to test for
differences between WT and CompoundX.

```{r}

dds <- DESeqDataSetFromMatrix(countData = round(count_data),
                              colData   = col_data,
                              design    = ~ condition)

# Run the full DESeq2 pipeline:
#   estimation of size factors, dispersion, and model fitting
dds <- DESeq(dds)

```

## 3 — PCA plot

A Principal Component Analysis gives us a quick overview of how the
samples cluster. We expect the WT and CompoundX samples to separate
along one of the principal components.

We use the **variance-stabilising transformation** (`vst`) since the dispersion
(variability) depends on the mean gene expression, which would bias the PCA.

```{r}

# Normally, you'd use (for many genes)
# vsd <- vst(dds, blind = FALSE)
# Now, we'll use:
vsd <- varianceStabilizingTransformation(dds, blind = FALSE)

plotPCA(vsd, intgroup = "condition") +
    theme_minimal() +
    ggtitle("PCA — WT vs CompoundX")

```

## 4 — Dispersion estimates

DESeq2 models each gene's dispersion (variability) as a
function of its mean expression. 
It is generally observed in RNA-seq data that the dispersion changes 
with different mean values.
The plot below shows the trend 
in the "dispersion estimate". The `vst()` function (or it's equivalent for very little data
 `varianceStabilizingTransformation()`) tries to estimate and counter
that trend.


```{r}

# What DESeq2 does
plotDispEsts(dds, main = "Dispersion estimates")
# The actual dispersion that we generated the data with
abline(h = 1/100, lty = 2, col = "grey")

```

DESeq2 uses dispersion estimates (determined for each gene) in two ways;

- When we use `plotPCA(vst(dds))` to generate a PCA plot, the PCA will work with a count table where
variance is stabilised across genes, such that each gene contributes equally to the PCA.
- Later, when statistical testing is performed, the dispersion is taken along
in the model fitting, which is important for accurate p-values and fold-change estimates.

::: {.callout-note}

## The difference between dispersion and variance

The dispersion estimate in DESeq2 is somewhat complicated. 
Technically, they assume:
$$
\text{var}(K) = \mu + \alpha \cdot \mu^2
$$

With K (multiple observations of) counts of a gene, $\mu$ the gene mean and $\alpha$ the gene's "dispersion estimate".

This is based on the assumption that the counts follow a Negative Binomial distribution, for which
this relationship between variance and the mean exists.

The dispersion estimate can thus be calculated as follows
$$
\alpha = \frac{\text{var}(K) - \mu}{\mu^2}
$$

Which is the `dispersion` value that is shown in the plot from `plotDispEsts()`. 

::: 

::: {.callout-note}

### How well does this model fit our generated data?

In our previous section, we generated data with a specific dispersion coefficient,
$\alpha=1/100$. So our generated data is a-typical, all genes should have the 
same dispersion coeffient $\alpha$, independent of their mean values.

I've added the grey dotted line to show that in the plot above. 
We can see that the data (black dots) indeed adhere that line.
However, DESeq2 thinks that the relationship (red dots) shows higher 
dispersion values for low-expressed genes. 
So for our limited generated data, it does **not** do a fantastic job.
This will work pretty well for real data though.

:::

## 5 — Results table

DESeq2 will fit our design (a formula describing the expected effect of our
condition) to each gene, 
taking into account the normalization (size factors) and the dispersion estimates,
and perform a Wald statistical test to determine 
whether the observed differences between conditions are significant. 
In fact, this was already done when we executed `dds <- DESeq(dds)`.

We can now extract the results for the contrast **CompoundX vs WT**. 
Key columns are:

| Column           | Meaning                                      |
|------------------|----------------------------------------------|
| `log2FoldChange` | log₂ fold-change (CompoundX / WT)            |
| `pvalue`         | raw p-value from the Wald test               |
| `padj`           | Benjamini–Hochberg adjusted p-value           |

```{r}

res <- results(dds, contrast = c("condition", "CompoundX", "WT"))

# Order by adjusted p-value
res_ordered <- res[order(res$padj), ]

# Show the top DEGs
head(as.data.frame(res_ordered), 20)

# How many genes pass padj < 0.05?
summary(res, alpha = 0.05)

```

The `padj` p value is the value you'll want to look at (as addressed
in the questions below). 

## 6 — Volcano plot

A volcano plot shows **statistical significance** (−log₁₀ adjusted
p-value) on the y-axis against **effect size** (log₂ fold-change) on the
x-axis. Genes that are both statistically significant and have a large
fold-change appear in the top-left and top-right corners.

```{r}

# Convert results to a data frame for ggplot
df_volcano <- as.data.frame(res) %>%
    rownames_to_column("Gene") %>%
    mutate(
        significant = case_when(
            is.na(padj)            ~ "NS",
            padj < 0.05 & log2FoldChange >  1 ~ "Up",
            padj < 0.05 & log2FoldChange < -1 ~ "Down",
            TRUE                               ~ "NS"
        )
    )

# Top 5 up-regulated and top 5 down-regulated genes (by padj)
df_volcano_top <- df_volcano %>%
    filter(significant %in% c("Up", "Down")) %>%
    group_by(significant) %>%
    slice_min(padj, n = 5) %>%
    ungroup()

# Make the plot
ggplot(df_volcano, aes(x = log2FoldChange, y = -log10(pvalue), colour = significant)) +
    geom_point(alpha = 0.7, size = 1.5) +
    scale_colour_manual(values = c("Up" = "firebrick", "Down" = "steelblue", "NS" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
    geom_vline(xintercept = c(-1, 1),     linetype = "dashed", colour = "grey40") +
    geom_text_repel(data = df_volcano_top, aes(label = Gene),
                    size = 3, max.overlaps = 20, show.legend = FALSE) +
    theme_minimal() +
    labs(title = "Volcano plot — CompoundX vs WT",
         x = expression(log[2]~fold~change),
         y = expression(-log[10]~p~value),
         colour = "Direction")

```

## Questions

We used generated data for this tutorial. These were the details we put in:

| Gene   | WT_reads | CompoundX_reads | Regulation   |
|--------|----------|-----------------|--------------|
| aceA   | 500      | 1200            | Up           |
| pykF   | 800      | 1600            | Up           |
| gltA   | 400      | 410             | Up           |
| crs    | 3        | 5               | Up           |
| tpiA   | 3        | 1000            | Up           |
| icd    | 1500     | 700             | Down         |
| pgi    | 900      | 400             | Down         |
| fumA   | 600      | 200             | Down         |
| lpdA   | 300      | 100             | Down         |
| zwf    | 20       | 15              | Down         |
| nrsX   | 1..1000* | idem            | None         |

_*)  1..1000 is a random number between 1 and 1000, equal for both conditions._

### Now answer the following questions

- Look at the results table, `res_ordered`:
    - When we look at the p-values column, ie `pvalue`, we see many genes with
    significant p-values. How come many of the "nrs" genes are also considered
    significant when we look at `pvalue`? 
        - What is p-value correction?
        - Does this succeed completely in its aim?
    - Why are not all genes that we know are DE not included in this list? (E.g. our 
    key transcription factor crs.)
    - What is our false discovery rate? And what is our power?
- Look at the Volcano plot:
    - Can you link each of the genes in the table with true values above
    to the Volcano plot? 
- Select all significant genes, for example `padj<0.05` from `res_ordered`. 
Can you explain this?
- Select a gene, and make a plot of its expression values, perhaps look into some
genes that show behavior you expected and did not expect (if any).



