---
title: "DESeq2 Deep Dive"
format: html
editor: visual
---

# Setting up

## Acknowledgements

Aside from the [DESeq2 paper itself](http://doi.org/10.1186/s13059-014-0550-8), additional [explanation](https://hbctraining.github.io/Intro-to-DGE/schedule/links-to-lessons.html) from [Mistry et al.](https://doi.org/10.5281/zenodo.4783480) from the Harvard Chan Bioinformatics Core was very helpfull in compiling this tutorial.

In addition, I used Claude Opus 4.6's understanding of DESeq2, and checked 
things in the original [DESeq2 source code](https://github.com/thelovelab/DESeq2) (15f2ec9).

```{=html}
<!--
Meeta Mistry, Mary Piper, Jihe Liu, & Radhika Khetani. (2021, May 24). hbctraining/DGE_workshop_salmon_online: Differential Gene Expression Workshop Lessons from HCBC (first release). Zenodo. https://doi.org/10.5281/zenodo.4783481. RRID:SCR_025373.
-->
```

## Libraries

Import libraries that we'll need.

```{r}
#| warning: false

library(tidyverse)
library(DESeq2)
library(pheatmap)
library(ggrepel)
library(scales)

# Set up output directory
if (!dir.exists("plots/tut-3")) {
    dir.create("plots/tut-3", recursive = TRUE, showWarnings = FALSE)
}

```

Define some functions and settings to make our plots look fancy later.

```{r}

# Bars should be black per default
update_geom_defaults("bar", list(fill = "black"))  # sets it globally for session

# Bigger arial font 
custom_adjust_fonts <- function(size = 15, family = "Arial") {
  theme(
    text             = element_text(size = size, family = family),
    axis.text        = element_text(size = size, family = family),
    axis.title       = element_text(size = size, family = family),
    plot.title       = element_text(size = size, family = family),
    plot.subtitle    = element_text(size = size, family = family),
    legend.text      = element_text(size = size, family = family),
    legend.title     = element_text(size = size, family = family),
    strip.text       = element_text(size = size, family = family)
  )
}

# Some bar graph settings I prefer
custom_theme_bar <- function() {
  list(
    theme_classic(),
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust=.5),    
      plot.title = element_text(hjust = .5)
    ),
    scale_y_continuous(
      labels = scales::label_comma(), 
      expand = expansion(mult = c(0, 0.05)),
      limits = c(0, NA)
      ),
    custom_adjust_fonts()
  )
}

```

## Load data

Let's first load two datasets: - The synthetic one that was generated in tutorial 1. - Example dataset [GLDS38](https://osdr.nasa.gov/bio/repo/data/studies/OSD-38), for which pre-processed count data is available in the `data-real/` directory.

```{r}
# Import the table
count_data_synth <- read.table("data_frozen/count_table.tsv", 
                         sep = "\t", 
                         header = TRUE, 
                         row.names = 1, 
                         check.names = FALSE)
# Also import real data (spaceflight)
# (Downloaded from https://github.com/mishapaauw/RNA-seq-workshop/tree/main/data
# where Misha created a downsampled version of the original data.)
count_data_real <- read.csv("/Users/m.wehrens/Data_UVA/example-datasets/2026_RNAseq-Misha/GLDS38_raw_counts.csv",
                            row.names=1)
count_data_real_meta <- read.csv("/Users/m.wehrens/Data_UVA/example-datasets/2026_RNAseq-Misha/samples_to_condition.csv",
                          row.names=1)
                          
# Now show 
```

And let's convert this to the tidy (aka 'long') format.

```{r}
# Convert synthetic data to long format
count_data_synth_long <- count_data_synth %>%
    rownames_to_column(var="gene") %>%
    pivot_longer(-gene, 
                names_to = "condition",
                values_to = "counts") 
    
# Convert real data to long format
count_data_real_long <- count_data_real %>%
    rownames_to_column(var="gene") %>%
    pivot_longer(-gene, 
                names_to = "sample",
                values_to = "counts") %>%

    # Additionally, set condition to FLT or GC, based on presence of _FLT_ or 
    # _GC_ string in sample name column
    mutate(condition = case_when(
        str_detect(sample, "_FLT_") ~ "FLT",
        str_detect(sample, "_GC_") ~ "GC",
        TRUE ~ sample
    )) |>
    
    # And create abbreviated sample names
    mutate(sample_brief=str_remove(sample, "Atha_WT_Col_0_sl_")) |>
    mutate(sample_brief=str_remove(sample_brief, "_G.S.$")) 
```

# Overview of the DESeq2 appraoch

When running the DESeq2 analysis, we see the following output:

```         
estimating size factors
estimating dispersions
gene-wise dispersion estimates
mean-dispersion relationship
final dispersion estimates
fitting model and testing
```

XXXXXXXXX

What happens during the DESeq2 analysis?

1.  **Normalization using size factors**
2.  **Dispersion estimate**
3.  **Estimate dispersion per gene**
4.  \*\*

# Normalization: size factors

If sample 1 in total has 5000 reads, and sample 2 in total has 10000 reads, and we consider a gene X, we expect that gene to have more reads in sample 2. In fact, a good guess would be that gene X has 2x as many reads in sample 2.

So, to more honestly compare samples, we need to *normalize* for the amount of reads in each samples.

When we have count table where rows correspond to genes, and columns to samples, we could simply divide each column by its total, and work with those values. Or, one step better, do that, and multiply with the mean of total sample counts, such that the values we end up with are in the same range as those with which we started.

DESeq2 takes a little more sophisticated appraoch. The total counts in a sample can be skewed by gene composition, e.g. a few highly expressed differentially expressed (DE) genes can affect total counts \[[1](https://hbctraining.github.io/Intro-to-DGE/lessons/02_DGE_count_normalization.html)\], \[[2](https://link.springer.com/article/10.1186/gb-2010-11-10-r106)\]. It therefor calculates a pseudo-reference sample, with arithmetic averages over the samples for each gene, then normalizes each gene by the reference sample, and then calculates the median normalized expression per sample. This somewhat complex scheme is summarized below:

| Gene | sampleA | sampleB | Pseudo-reference sample | sampleA/ref | sampleB/ref |
|------------|------------|------------|----------------|------------|------------|
| EF2A | 1489 | 906 | sqrt(1489 \* 906) = 1161.5 | 1489/1161.5 = 1.28 | 906/1161.5 = 0.78 |
| ABCD1 | 22 | 13 | sqrt(22 \* 13) = 17.7 | 22/16.9 = 1.30 | 13/16.9 = 0.77 |
| ... | ... | ... | ... | ... | ... |
|  |  |  |  | **size factor sampleA:** median(1.28, 1.30, ..) = 1.30 | **size factor sampleB:** median(0.78, 0.77, ..) = 0.78 |

Let's go over this with some code.

## Total reads per sample

First, let's simply show the total reads per sample:

```{r}
# First determine total reads per sample
df_total_reads <- count_data_real_long |>
    group_by(sample) |>
    summarize(total_counts = sum(counts))
    
# Now plot in bar graph
p <- df_total_reads |>
  mutate(sample=str_remove(sample, "Atha_WT_Col_0_sl_")) |>
  ggplot(aes(x=sample, y=total_counts)) +
    geom_bar(stat="identity") +
    xlab('Sample')+ylab('Total counts')+
    custom_theme_bar() 
  
# save
ggsave("plots/tut-3/total_counts_per_sample.pdf", plot=p, width=10, height=16, units='cm', device=cairo_pdf)
# and show
p + custom_adjust_fonts(12)
```

## Estimate size factors manually

Now, estimate the size factors per sample, the same way DESeq2 does it.

```{r}

# Estimate the "size factors"
# Create reference sample; geometric mean
df_size_factors <- count_data_real_long |>  
  # Calculate geometric means of gene values
  group_by(gene) |>
  mutate(geom_mean = exp(mean(log(counts)))) |>
    # More explicit alternative; theoretically same, though less numerically stable
    # mutate(geom_mean = median(prod(counts)^(1/(length(counts))))) |>
  ungroup() |>
  # Filter out zero values (choice following DESeq2)
  filter(geom_mean > 0) |>
  # Now normalize the counts by geom mean
  mutate(counts_norm = counts/geom_mean) |>
  # And calculate median normalized counts (=size factor)
  group_by(sample) |>
  summarize(size_factor = median(counts_norm, na.rm=TRUE))

# now plot those
p <- df_size_factors |>
  mutate(sample=str_remove(sample, "Atha_WT_Col_0_sl_")) |>
  ggplot(aes(x=sample, y=size_factor)) +
  geom_bar(stat="identity") +
  xlab('Sample') + ylab('Size factor')
  custom_theme_bar()
# save
ggsave("plots/tut-3/size_factors_per_sample.pdf", plot=p, width=10, height=16, units='cm', device=cairo_pdf)
# show
p + custom_adjust_fonts(12)

```

## Compare manual size factors with DESeq2

As a sanity check, let's see whether this complies with DESeq2:

```{r}

# Let's see how well this matches DESeq2's assessment

# Set up a dds
dds <- DESeqDataSetFromMatrix(countData = count_data_real,
                              colData   = count_data_real_meta,
                              design    = ~ condition)
dds <- estimateSizeFactors(dds)
the_sizefactors <- sizeFactors(dds)
df_size_factors_deseq2 <- data.frame(
  size_factor = unname(the_sizefactors),
  sample = names(the_sizefactors))

# now comparison
df_size_factors_all <-  
  rbind( df_size_factors |> mutate(origin="manual"), 
        df_size_factors_deseq2 |> mutate(origin = "deseq2"))
        
# now plot
p <- df_size_factors_all |>
  mutate(sample = str_remove(sample, "Atha_WT_Col_0_sl_")) |>
  ggplot(aes(x = sample, y = size_factor, fill = origin)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_manual(values = c("manual" = "black", "deseq2" = "red")) + 
  xlab('Sample') + ylab('Size factor') +
  custom_theme_bar()
  
ggsave("plots/tut-3/size_factors_manual_vs_deseq2.pdf", plot = p, width = 10, height = 16, units = "cm", device=cairo_pdf)

p + custom_adjust_fonts(12)
```

## Manually normalize genes and see effect

Let's now apply normalization of the count values, and illustrate the effect with a plot for 10 random genes.

```{r}

# set seed for reproducibility
set.seed(123)

# add size factors to count table, and add normalized counts to table 
count_data_real_long_normalized <- count_data_real_long |>
  left_join(df_size_factors, by='sample') |>
  mutate(normalized_count = counts/size_factor)

# plot 10 random genes, with raw and normalized counts
p <- count_data_real_long_normalized |> 
  pivot_longer(cols=c(counts, normalized_count), names_to="type", values_to = "counts") |>
  filter(gene %in% sample(unique(gene), 9)) |>  
  ggplot(aes(x=sample_brief, y=counts, color=type, shape=condition)) +
    # geom_bar(stat="identity", position="dodge") +
    geom_point() +
    facet_wrap(~gene, scales = "free_y", ncol = 3) +
    theme_classic() +
    xlab('Condition') + ylab('Counts') +
    scale_color_manual(values = c("counts" = "black", "normalized_count" = "red")) + 
    # scale_color_manual(values = c("FLT" = "darkblue", "GC" = "brown")) +
    custom_theme_bar()

# and save
ggsave("plots/tut-3/normalized_counts_example_raw.pdf", plot=p, width=20, height=15, units='cm', device=cairo_pdf)

p + custom_adjust_fonts(12)

```

In the plot above, the red points better reflect the true transcript concentrations compared between the samples.

As a final point, note that the total counts for each sample are not after DESeq2's normalization.

```{r}
# Let's do a full run
dds_full <- DESeq(dds)

# We can now also get normalized count matrix from deseq2
normalized_counts <- counts(dds_full, normalized=TRUE)
# Which doesn't have all sample totals equal
colSums(normalized_counts)


p <- count_data_real_long_normalized |> 
  pivot_longer(cols=c(counts, normalized_count), names_to="type", values_to = "counts") |>
  group_by(sample_brief, type) |>
  summarize(total_counts = sum(counts)) |>
  ungroup() |> 
  ggplot(aes(x=sample_brief, y=total_counts, fill=type)) +
    geom_bar(stat="identity", position="dodge") +
    theme_classic() +
    xlab('Condition') + ylab('Total counts') +
    scale_fill_manual(values = c("counts" = "grey", "normalized_count" = "black")) + 
    custom_theme_bar()

ggsave(
  "plots/tut-3/normalization_comparison.pdf", 
  plot=p, width=20, height=15, units='cm', device=cairo_pdf
  )

```

# Calculate dispersions

## What's dispersion and why is it important

In the end, our aim is to know how changes in gene expression depend on the different conditions of our experiment.

Due to technical and biological factors, gene expression counts will vary, even within the same condition.
Dispersion (variability) refers to the extent by which observations of a quantity, in this case counts, vary.

In the example with gene A below, we can say with high likelyhood that the gene is differentially expressed, whilst 
in the example with gene B, we can be less sure of that.

```{r}
set.seed(30)

df_example_simpleDE <-
  data.frame(
    counts = c(
      20+rnorm(n = 3, mean = 10, sd = .25),
      20+rnorm(n = 3, mean = 15, sd = .25),
      20+rnorm(n = 3, mean = 10, sd = 15),
      20+rnorm(n = 3, mean = 15, sd = 15)
    ),
    gene = c(
      rep('gene A\n(low dispersion)', 6),
      rep('gene B\n(high dispersion)', 6)
    ),
    condition = c(
      rep('Control',3), 
      rep('Treatment',3), 
      rep('Control',3), 
      rep('Treatment',3)
    ))
    
p <- df_example_simpleDE |>
  ggplot(aes(x=condition, y=counts))+
  geom_point()+
  facet_grid(cols = vars(gene))+
  theme_classic() +
  ylab("Counts")+
  xlab("Condition")+
  ggtitle("n=3") +
  theme(plot.title = element_text(hjust = 1)) +
  custom_adjust_fonts()
  
# save 
ggsave("plots/tut-3/dispersion_example.pdf", plot=p, width=13, height=12, units='cm', device=cairo_pdf)

p + custom_adjust_fonts(12)
```

Simply put, DESeq2 will try to estimate the dispersion in each gene's counts, and use that 
estimate to perform statistics and estimate the likelyhood that there is differential expression
between conditions. 
Or, in more statistically sound terms, it will calculate the p-value based on those estimates, 
which reflects the likelyhood that H0 (no difference between conditions),
is true. If that probability is low, we can reject H0, and say that there is likely a difference between conditions.

The method by which this is done is quite a bit more complicated, but that's the 
gist of it.

### The dispersion is not similar for all genes

Let's take a better look at the dispersion, and calculate some metrics that
quantify that dispersion.

```{r}

# First for the synthetic data
# now calculate mean, variance and sd
df_dispersion_synth <- count_data_synth_long %>%      
    group_by(gene) %>%
    summarise(mean = mean(counts),
              sd   = sd(counts),
              variance = var(counts)) %>%
    # include "alpha", which we'll use later
    mutate(alpha = (variance-mean)/mean^2)
    
# Then for the real data
# now calculate mean, variance and sd    
df_dispersion_real <- count_data_real_long %>%      
    group_by(gene) %>%
    summarise(mean = mean(counts),
              sd   = sd(counts),
              variance = var(counts)) %>%
    # include "alpha", which we'll use later
    mutate(alpha = (variance-mean)/mean^2)
```

When we look at the values that we now have, we'll see that there is a wide
range of dispersion values per gene:

```{r}
# Histogram of dispersion values for all genes

p <- df_dispersion_real |>
    ggplot(aes(x=sd)) +
    geom_histogram(bins=100) +
    theme_classic() +
    scale_x_log10(labels = scales::label_comma()) +
    xlab("Standard deviation") +
    ylab("Times observed") +
    ggtitle("Spaceflight data") +
    custom_theme_bar()
    
ggsave("plots/tut-3/dispersion_histogram.pdf", plot=p, width=10, height=10, units='cm', device=cairo_pdf)

p + custom_adjust_fonts(12)

```

## DESeq2's handling of dispersion

Estimating dispersion (variabillity) play a big role in DESeq2's approach. Four of DESeq2's messages refer to this process.

```         
estimating dispersions
gene-wise dispersion estimates
mean-dispersion relationship
final dispersion estimates
```

The first output announces the start of the dispersion section[^Explanation Claude, and see methods.R (`estimateDispersions.DESeqDataSet <- function(...`), core.R in DESeq2 [source](https://github.com/thelovelab/DESeq2) 15f2ec9.], the 
three that follow refer to sub-steps of that procedure.

I don't pretend to understand the DESeq2 code in full, but we can emulate 
the process that DESeq2 follows to understand what's going on.







```{r}
plotDispEsts(dds_full)
```


```{r}
# Information about the dispersion of gene counts plays a big role in
# the DESeq2 analysis.
#
# Dispersion (~variability) can be quantified in multiple way.
# 
# Here, we calculate the sd and mean, and additionally "alpha".
# Alpha is (variance-mean)/mean^2).

# this script yields
# - df_dispersion_synth
# - df_dispersion_real
```



# Variance, mean, and dispersion

## Variance versus mean

```{r}
# DESeq2 has a dispersion estimate parameter, alpha, defined by
# var(K) = mu + alpha * mu^2
# Thus
# alpha = (var(K)-mu) / mu^2
              
# Now plot variance versus mean
#
# Very nice source of information:
# https://hbctraining.github.io/DGE_workshop/lessons/01_DGE_setup_and_overview.html#rna-seq-count-distribution
# The red line gives the poisson relationship of the mean and variance
# The blue line gives the negative binominal relationship between mean and variance
fn_NBvariance <- function(mu, alpha) {
    return(mu + alpha * mu^2)
}
MY_ALPHA = 1/100
ggplot(df_dispersion_synth, aes(x=mean, y=variance)) +
    geom_point() + 
    geom_abline(slope=1, intercept=0, color='red') +
    stat_function(fun=fn_NBvariance, args=list(alpha=MY_ALPHA), color='blue') +
    scale_x_log10() + scale_y_log10() + 
    theme_bw() +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)')
```

## Dispersion estimate plot

```{r}
# Now plot (var(K)-mu) / mu^2 versus mean
# This produces a similar plot to the dispresion estimate plot of DESeq2
ggplot(df_dispersion_synth, aes(x=mean, y=alpha)) +
    geom_point() +
    scale_x_log10() + scale_y_log10(limits=c(1e-8, 10))+
    xlab("mean") + ylab("dispersion estimate\nalpha = (var(K)-mu) / mu^2")

# Note furthermore that DESeq2 selects the top-variance genes for taking 
# along in the PCA. 
```

## Fit dispersion on real data

```{r}
# Let's first fit the dispersion coefficient
# Fit alpha in: variance = mu + alpha * mu^2
# Use nls on log-transformed values for a better fit across the wide range
df_fit <- df_dispersion_real %>% filter(mean > 0, variance > 0)
fit <- nls(log(variance) ~ log(mean + alpha * mean^2),
           data = df_fit,
           start = list(alpha = 0.01),
           control = nls.control(maxiter = 200))
MY_ALPHA_real <- coef(fit)[["alpha"]]
cat("Fitted alpha:", MY_ALPHA_real, "\n")
# 
p <- ggplot(df_dispersion_real, aes(x=mean, y=variance)) +
    geom_point(color='grey') + 
    geom_abline(slope=1, intercept=0, color='red', size=1) +
    stat_function(fun=fn_NBvariance, args=list(alpha=MY_ALPHA_real), color='blue', size=1) +
    scale_x_log10() + scale_y_log10() + 
    theme_classic() +
    annotate("text", x = 0.1, y = max(df_dispersion_real$variance)*.95, 
             label = paste0("variance = mean + alpha * mean^2\nalpha =",round(MY_ALPHA,2)),
             hjust = 0, vjust = 1) +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)') +
    ggtitle('Spaceflight data')
# p
# export to directory plots/tut-3/
ggsave("plots/tut-3/variance_mean_real_data.pdf", plot=p, width=10, height=10, units='cm')
```

# Log2 fold change

```{r}
# Calculate means and log2fc per condition
count_data_real_long_l2fc <- count_data_real_long |>
    # first gather averages
    group_by(condition, gene) |>
    summarise(gene_average = mean(counts)) |>    
    # now determine FC
    filter(gene_average>0) |>
    group_by(gene) |>
    pivot_wider(names_from = condition, values_from = gene_average) |>
    mutate(FC = FLT/GC, 
            L2FC = log2(FC)) |>
    arrange(desc(L2FC))
```

```{r}
# now make a small plot for gene="AT2G41355"
MYGENE = "AT2G41355"
p <- count_data_real_long |> 
    filter(gene==MYGENE) |>
    ggplot(aes(x=condition, y=counts))+
        geom_jitter(height = 0, width = 0.1) +
        theme_classic() +
        ggtitle(MYGENE)
ggsave("plots/tut-3/gene_example_highl2fc.pdf", plot=p, width=7, height=7, units='cm')
```

# PCA fundamentals

## PCA on synthetic count data

```{r}
# PCA expects samples as rows, genes as columns, so we transpose.
# We also scale and center the data (z-scores), which is standard practice.

pca_result <- prcomp(t(count_data_synth), center = TRUE, scale. = TRUE)

# How much variance does each PC explain?
pca_var <- summary(pca_result)$importance
print(pca_var[, 1:4])
```

## Scree plot

```{r}
# Plot: variance explained per PC
df_var_explained <- data.frame(
    PC = paste0("PC", 1:ncol(pca_var)),
    variance_explained = pca_var["Proportion of Variance", ]
) |> mutate(PC = factor(PC, levels = PC))

p_scree <- ggplot(df_var_explained, aes(x = PC, y = variance_explained)) +
    geom_bar(stat = "identity") +
    theme_classic() +
    ylab("Proportion of variance explained") +
    ggtitle("Scree plot - synthetic data")
p_scree
```

## PCA scatter (PC1 vs PC2)

```{r}
df_scores <- as.data.frame(pca_result$x) |>
    mutate(sample = rownames(pca_result$x),
           condition = ifelse(str_detect(sample, "^WT"), "WT", "CompoundX"))

p_pca <- ggplot(df_scores, aes(x = PC1, y = PC2, color = condition)) +
    geom_point(size = 3) +
    theme_classic() +
    xlab(paste0("PC1 (", round(pca_var["Proportion of Variance", 1] * 100, 1), "%)")) +
    ylab(paste0("PC2 (", round(pca_var["Proportion of Variance", 2] * 100, 1), "%)")) +
    ggtitle("PCA - synthetic data")
p_pca
```

## Gene loadings

```{r}
# The rotation matrix contains the loadings (weights) per gene per PC.

df_loadings <- as.data.frame(pca_result$rotation) |>
    rownames_to_column("gene") |>
    mutate(PC1_abs = abs(PC1), PC2_abs = abs(PC2))

# Top 15 genes by PC1 loading
top_pc1 <- df_loadings |> arrange(desc(PC1_abs)) |> head(15)
print("Top genes contributing to PC1:")
print(top_pc1 |> select(gene, PC1))

# Top 15 genes by PC2 loading
top_pc2 <- df_loadings |> arrange(desc(PC2_abs)) |> head(15)
print("Top genes contributing to PC2:")
print(top_pc2 |> select(gene, PC2))

# Plot loadings for PC1
p_loadings <- top_pc1 |>
    mutate(gene = fct_reorder(gene, PC1)) |>
    ggplot(aes(x = gene, y = PC1)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_classic() +
    ggtitle("Top 15 gene loadings on PC1")
p_loadings
```

## tpiA loading across all PCs

```{r}
df_tpiA <- df_loadings |>
    filter(gene == "tpiA") |>
    select(gene, starts_with("PC"), -PC1_abs, -PC2_abs) |>
    pivot_longer(-gene, names_to = "PC", values_to = "loading") |>
    mutate(PC = factor(PC, levels = paste0("PC", 1:8)))

p_tpiA <- ggplot(df_tpiA, aes(x = PC, y = loading)) +
    geom_bar(stat = "identity") +
    theme_classic() +
    ggtitle("tpiA loading across all PCs")
p_tpiA
```

# PCA on normalized data

```{r}
# PCA on normalized (variance-stabilized) data
# This builds on the normalization from the size factors section,
# re-creating the DESeq2 object to then apply VST and run PCA.
```

## Variance vs mean (raw)

```{r}
# There tends to be a trend in the variance, compared to mean expression values.
p <- ggplot(df_dispersion_real, aes(x=mean, y=variance)) +
    geom_point(color='grey') + 
    scale_x_log10() + scale_y_log10() + 
    theme_classic() +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)') +
    ggtitle('Spaceflight data')
p
ggsave("plots/tut-3/variance_mean_real_data_raw.pdf", plot=p, width=10, height=10, units='cm')
```

## Z-score normalization

```{r}
# If we were to use z-scores
# (Often done when using PCAs)
count_data_real_long_zscore <- count_data_real_long |>
  group_by(gene) |>
  mutate(counts_z = counts/sd(counts)) |>
  ungroup()
# Plot our gene
MYGENE = "AT2G41355"
p <- count_data_real_long_zscore |> 
    filter(gene==MYGENE) |>
    ggplot(aes(x=condition, y=counts_z))+
        geom_jitter(height = 0, width = 0.1) +
        theme_classic() +
        ggtitle(MYGENE)
p
```

```{r}
# now calculate dispersions z
df_dispersion_real_zscore <- count_data_real_long_zscore |>
    group_by(gene) |>
    summarise(mean = mean(counts_z),
              sd   = sd(counts_z),
              variance = var(counts_z)) |>
    mutate(alpha = (variance-mean)/mean^2)
# now plot dispersions z
p <- ggplot(df_dispersion_real_zscore, aes(x=mean, y=variance)) +
    geom_point(color='grey') + 
    scale_x_log10() + scale_y_log10() + 
    theme_classic() +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)') +
    ggtitle('Spaceflight data, z-scores')
p
ggsave("plots/tut-3/variance_mean_real_data_zscore.pdf", plot=p, width=10, height=10, units='cm')
```

```{r}
# similarly, we'd see a trend if we'd take the log
df_disperson_log <- count_data_real_long_zscore |>
    mutate(      
      log_counts   = log(counts+.1),
      log_counts_z = log(counts_z+.1)) |>
    group_by(gene) |>
    summarise(
      mean_z = mean(log_counts_z),
      sd_z   = sd(log_counts_z),
      variance_z = var(log_counts_z),
      mean = mean(log_counts), 
      sd = sd(log_counts),
      variance = var(log_counts))
# plot again
p <- ggplot(df_disperson_log, aes(x=mean, y=variance)) +
    geom_point(color='grey') + 
    scale_x_log10() + scale_y_log10() + 
    theme_classic() +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)') +
    ggtitle('Spaceflight data')
p
```

## Variance-stabilizing transformation (VST)

```{r}
# So now let's check out how dispersions look using vst
vsd <- vst(dds)
vst_counts <- assay(vsd)
df_dispersion_ddsvst <- vst_counts |>
    as.data.frame() |>
    rownames_to_column("gene") |>
    pivot_longer(-gene, names_to = "sample", values_to = "counts") |>
    group_by(gene) |>
    summarise(mean = mean(counts),
              sd   = sd(counts),
              variance = var(counts)) |>
    mutate(alpha = (variance - mean) / mean^2)
# now plot
p <- ggplot(df_dispersion_ddsvst, aes(x=mean, y=variance)) +
    geom_point(color='grey') + 
    scale_x_log10() + scale_y_log10() + 
    theme_classic() +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)') +
    ggtitle('Spaceflight data, variance stabilized')
p
ggsave("plots/tut-3/variance_mean_real_data_vst.pdf", plot=p, width=10, height=10, units='cm')
```

## PCA on VST data

```{r}
# now a pca plot by deseq2
pca_res <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
p <- ggplot(pca_res, aes(x=PC1, y=PC2, shape=condition,
                      color=condition)) +
    geom_point(size=3) +
    theme_classic() +
    xlab(paste0("PC1 (", round(100*pca_res$percentVar[1], 1), "%)")) +
    ylab(paste0("PC2 (", round(100*pca_res$percentVar[2], 1), "%)")) +
    ggtitle("PCA on vst data") +
    scale_color_manual(values=c("space_flight"="darkblue", "ground_control"="brown")) +
    theme(legend.position = "none")
p
ggsave("plots/tut-3/pca_deseq2.pdf", plot=p, width=10, height=10, units='cm')
```

## Normalized count totals (recap)

```{r}
# Also in the overall DESeq2 analysis, dispersion-stablization is used
# (See also https://hbctraining.github.io/Intro-to-DGE/lessons/04b_DGE_DESeq2_analysis.html)

# So we have
# - Size factor correction
# - Dispersion correction

# Plot, remind ourselves that also the per-sample count normalization went well
normalized_counts_sampletotals <- colSums(normalized_counts)
p <- 
  data.frame(
    counts = unname(normalized_counts_sampletotals),
    sample = names(normalized_counts_sampletotals)
  ) |>
  mutate(sample = str_remove(sample, "Atha_WT_Col_0_sl_")) |>
  ggplot(aes(x = sample, y = counts)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("Sample") + ylab("Counts\n(normalized by size factors)")
ggsave("plots/tut-3/total_counts_per_sample_normalized.pdf", plot=p, width=7, height=10, units='cm')


```

# Statistical testing

```{r}
```