---
title: "DESeq2 Deep Dive"
format: html
editor: visual
---

# Load data

```{r}
# Loads two example datasets;
# - Synthetic one that was generated in main dir of this repo
# - Example dataset GLDS38, see https://osdr.nasa.gov/bio/repo/data/studies/OSD-38 (pre-processed to counts)

# this script yields
# - count_data_synth
# - count_data_real
# - count_data_synth_long
# - count_data_real_long
```

```{r}
# Import the table
count_data_synth <- read.table("data_frozen/count_table.tsv", 
                         sep = "\t", 
                         header = TRUE, 
                         row.names = 1, 
                         check.names = FALSE)
# Also import real data (spaceflight)
# (Downloaded from https://github.com/mishapaauw/RNA-seq-workshop/tree/main/data
# where Misha created a downsampled version of the original data.)
count_data_real <- read.csv("/Users/m.wehrens/Data_UVA/example-datasets/2026_RNAseq-Misha/GLDS38_raw_counts.csv",
                            row.names=1)
count_data_real_meta <- read.csv("/Users/m.wehrens/Data_UVA/example-datasets/2026_RNAseq-Misha/samples_to_condition.csv",
                          row.names=1)
print("Loaded count_data_synth and count_data_real")
```

```{r}
# make the format long
count_data_synth_long <- count_data_synth %>%
    rownames_to_column(var="gene") %>%
    pivot_longer(-gene, 
                names_to = "condition",
                values_to = "counts") 
    
count_data_real_long <- count_data_real %>%
    rownames_to_column(var="gene") %>%
    pivot_longer(-gene, 
                names_to = "sample",
                values_to = "counts") %>%
    # set condition to FLT or GC, based on presence of _FLT_ or _GC_ string in condtion column
    mutate(condition = case_when(
        str_detect(sample, "_FLT_") ~ "FLT",
        str_detect(sample, "_GC_") ~ "GC",
        TRUE ~ sample
    ))
print("Formatted data in long format, count_data_synth_long, count_data_real_long")
```

# Calculate dispersions

```{r}
# Information about the dispersion of gene counts plays a big role in
# the DESeq2 analysis.
#
# Dispersion (~variability) can be quantified in multiple way.
# 
# Here, we calculate the sd and mean, and additionally "alpha".
# Alpha is (variance-mean)/mean^2).

# this script yields
# - df_dispersion_synth
# - df_dispersion_real
```

```{r}
# now calculate mean, variance and sd
df_dispersion_synth <- count_data_synth_long %>%      
    group_by(gene) %>%
    summarise(mean = mean(counts),
              sd   = sd(counts),
              variance = var(counts)) %>%
    mutate(alpha = (variance-mean)/mean^2)
    
df_dispersion_real <- count_data_real_long %>%      
    group_by(gene) %>%
    summarise(mean = mean(counts),
              sd   = sd(counts),
              variance = var(counts)) %>%
    mutate(alpha = (variance-mean)/mean^2)
print("Calculated dispersion data frame, df_dispersion_synth, df_dispersion_real")
```

# Normalization: size factors

```{r}
library(tidyverse)
library(ggplot2)

library(DESeq2)
```

## Total reads per sample

```{r}
# First determine total reads per sample
df_total_reads <- count_data_real_long |>
    group_by(sample) |>
    summarize(total_counts = sum(counts))
    
# Now plot in bar graph
p <- df_total_reads |>
  mutate(sample=str_remove(sample, "Atha_WT_Col_0_sl_")) |>
  ggplot(aes(x=sample, y=total_counts)) +
    geom_bar(stat="identity") + 
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
# p    
# save
ggsave("plots/total_counts_per_sample.pdf", plot=p, width=7, height=7, units='cm')
```

## Estimate size factors manually

```{r}
# Estimate the "size factors"
# Create reference sample; geometric mean
df_size_factors <- count_data_real_long |>  
  # Calculate geometric means of gene values
  group_by(gene) |>
  mutate(geom_mean = median(prod(counts)^(1/(length(counts))))) |>
  ungroup() |>
  # Now normalize the counts by geom mean
  mutate(counts_norm = counts/geom_mean) |>
  # And calculate median normalized counts (=size factor)
  group_by(sample) |>
  summarize(size_factor = median(counts_norm, na.rm=TRUE))

# now plot those
p <- df_size_factors |>
  mutate(sample=str_remove(sample, "Atha_WT_Col_0_sl_")) |>
  ggplot(aes(x=sample, y=size_factor)) +
  geom_bar(stat="identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
# p    
# save
ggsave("plots/size_factors_per_sample.pdf", plot=p, width=7, height=7, units='cm')
```

## Compare manual size factors with DESeq2

```{r}
# Let's see how well this matches DESeq2's assessment
dds <- DESeqDataSetFromMatrix(countData = count_data_real,
                              colData   = count_data_real_meta,
                              design    = ~ condition)
dds <- estimateSizeFactors(dds)
the_sizefactors <- sizeFactors(dds)
df_size_factors_deseq2 <- data.frame(
  size_factor = unname(the_sizefactors),
  sample = names(the_sizefactors))

# now comparison
df_size_factors_all <-  
  rbind( df_size_factors |> mutate(origin="manual"), 
        df_size_factors_deseq2 |> mutate(origin = "deseq2"))
# now plot
p <- df_size_factors_all |>
  mutate(sample = str_remove(sample, "Atha_WT_Col_0_sl_")) |>
  ggplot(aes(x = sample, y = size_factor, fill = origin)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_manual(values = c("manual" = "black", "deseq2" = "red"))
# p
ggsave("plots/size_factors_manual_vs_deseq2.pdf", plot = p, width = 9, height = 7, units = "cm")

# We can now also get normalized count matrix from deseq2
normalized_counts <- counts(dds, normalized=TRUE)
# Which doesn't have all sample totals equal
colSums(normalized_counts)
```

## Normalized count totals and full DESeq2 analysis

```{r}
# Also in the overall DESeq2 analysis, dispersion-stablization is used
# (See also https://hbctraining.github.io/Intro-to-DGE/lessons/04b_DGE_DESeq2_analysis.html)

# So we have
# - Size factor correction
# - Dispersion correction

# Plot, remind ourselves that also the per-sample count normalization went well
normalized_counts_sampletotals <- colSums(normalized_counts)
p <- 
  data.frame(
    counts = unname(normalized_counts_sampletotals),
    sample = names(normalized_counts_sampletotals)
  ) |>
  mutate(sample = str_remove(sample, "Atha_WT_Col_0_sl_")) |>
  ggplot(aes(x = sample, y = counts)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("Sample") + ylab("Counts\n(normalized by size factors)")
ggsave("plots/total_counts_per_sample_normalized.pdf", plot=p, width=7, height=10, units='cm')


# Run the whole analysis
dds_2 <- DESeq(dds) # saving to dds_2 here, usually not done
```

# Variance, mean, and dispersion

## Variance versus mean

```{r}
# DESeq2 has a dispersion estimate parameter, alpha, defined by
# var(K) = mu + alpha * mu^2
# Thus
# alpha = (var(K)-mu) / mu^2
              
# Now plot variance versus mean
#
# Very nice source of information:
# https://hbctraining.github.io/DGE_workshop/lessons/01_DGE_setup_and_overview.html#rna-seq-count-distribution
# The red line gives the poisson relationship of the mean and variance
# The blue line gives the negative binominal relationship between mean and variance
fn_NBvariance <- function(mu, alpha) {
    return(mu + alpha * mu^2)
}
MY_ALPHA = 1/100
ggplot(df_dispersion, aes(x=mean, y=variance)) +
    geom_point() + 
    geom_abline(slope=1, intercept=0, color='red') +
    stat_function(fun=fn_NBvariance, args=list(alpha=MY_ALPHA), color='blue') +
    scale_x_log10() + scale_y_log10() + 
    theme_bw() +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)')
```

## Dispersion estimate plot

```{r}
# Now plot (var(K)-mu) / mu^2 versus mean
# This produces a similar plot to the dispresion estimate plot of DESeq2
ggplot(df_dispersion, aes(x=mean, y=alpha)) +
    geom_point() +
    scale_x_log10() + scale_y_log10(limits=c(1e-8, 10))+
    xlab("mean") + ylab("dispersion estimate\nalpha = (var(K)-mu) / mu^2")

# Note furthermore that DESeq2 selects the top-variance genes for taking 
# along in the PCA. 
```

## Fit dispersion on real data

```{r}
# Let's first fit the dispersion coefficient
# Fit alpha in: variance = mu + alpha * mu^2
# Use nls on log-transformed values for a better fit across the wide range
df_fit <- df_dispersion_real %>% filter(mean > 0, variance > 0)
fit <- nls(log(variance) ~ log(mean + alpha * mean^2),
           data = df_fit,
           start = list(alpha = 0.01),
           control = nls.control(maxiter = 200))
MY_ALPHA_real <- coef(fit)[["alpha"]]
cat("Fitted alpha:", MY_ALPHA_real, "\n")
# 
p <- ggplot(df_dispersion_real, aes(x=mean, y=variance)) +
    geom_point(color='grey') + 
    geom_abline(slope=1, intercept=0, color='red', size=1) +
    stat_function(fun=fn_NBvariance, args=list(alpha=MY_ALPHA_real), color='blue', size=1) +
    scale_x_log10() + scale_y_log10() + 
    theme_classic() +
    annotate("text", x = 0.1, y = max(df_dispersion_real$variance)*.95, 
             label = paste0("variance = mean + alpha * mean^2\nalpha =",round(MY_ALPHA,2)),
             hjust = 0, vjust = 1) +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)') +
    ggtitle('Spaceflight data')
# p
# export to directory plots/
ggsave("plots/variance_mean_real_data.pdf", plot=p, width=10, height=10, units='cm')
```

# Log2 fold change

```{r}
# Calculate means and log2fc per condition
count_data_real_long_l2fc <- count_data_real_long |>
    # first gather averages
    group_by(condition, gene) |>
    summarise(gene_average = mean(counts)) |>    
    # now determine FC
    filter(gene_average>0) |>
    group_by(gene) |>
    pivot_wider(names_from = condition, values_from = gene_average) |>
    mutate(FC = FLT/GC, 
            L2FC = log2(FC)) |>
    arrange(desc(L2FC))
```

```{r}
# now make a small plot for gene="AT2G41355"
MYGENE = "AT2G41355"
p <- count_data_real_long |> 
    filter(gene==MYGENE) |>
    ggplot(aes(x=condition, y=counts))+
        geom_jitter(height = 0, width = 0.1) +
        theme_classic() +
        ggtitle(MYGENE)
ggsave("plots/gene_example_highl2fc.pdf", plot=p, width=7, height=7, units='cm')
```

# PCA fundamentals

## PCA on synthetic count data

```{r}
# PCA expects samples as rows, genes as columns, so we transpose.
# We also scale and center the data (z-scores), which is standard practice.

pca_result <- prcomp(t(count_data_synth), center = TRUE, scale. = TRUE)

# How much variance does each PC explain?
pca_var <- summary(pca_result)$importance
print(pca_var[, 1:4])
```

## Scree plot

```{r}
# Plot: variance explained per PC
df_var_explained <- data.frame(
    PC = paste0("PC", 1:ncol(pca_var)),
    variance_explained = pca_var["Proportion of Variance", ]
) |> mutate(PC = factor(PC, levels = PC))

p_scree <- ggplot(df_var_explained, aes(x = PC, y = variance_explained)) +
    geom_bar(stat = "identity") +
    theme_classic() +
    ylab("Proportion of variance explained") +
    ggtitle("Scree plot - synthetic data")
p_scree
```

## PCA scatter (PC1 vs PC2)

```{r}
df_scores <- as.data.frame(pca_result$x) |>
    mutate(sample = rownames(pca_result$x),
           condition = ifelse(str_detect(sample, "^WT"), "WT", "CompoundX"))

p_pca <- ggplot(df_scores, aes(x = PC1, y = PC2, color = condition)) +
    geom_point(size = 3) +
    theme_classic() +
    xlab(paste0("PC1 (", round(pca_var["Proportion of Variance", 1] * 100, 1), "%)")) +
    ylab(paste0("PC2 (", round(pca_var["Proportion of Variance", 2] * 100, 1), "%)")) +
    ggtitle("PCA - synthetic data")
p_pca
```

## Gene loadings

```{r}
# The rotation matrix contains the loadings (weights) per gene per PC.

df_loadings <- as.data.frame(pca_result$rotation) |>
    rownames_to_column("gene") |>
    mutate(PC1_abs = abs(PC1), PC2_abs = abs(PC2))

# Top 15 genes by PC1 loading
top_pc1 <- df_loadings |> arrange(desc(PC1_abs)) |> head(15)
print("Top genes contributing to PC1:")
print(top_pc1 |> select(gene, PC1))

# Top 15 genes by PC2 loading
top_pc2 <- df_loadings |> arrange(desc(PC2_abs)) |> head(15)
print("Top genes contributing to PC2:")
print(top_pc2 |> select(gene, PC2))

# Plot loadings for PC1
p_loadings <- top_pc1 |>
    mutate(gene = fct_reorder(gene, PC1)) |>
    ggplot(aes(x = gene, y = PC1)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_classic() +
    ggtitle("Top 15 gene loadings on PC1")
p_loadings
```

## tpiA loading across all PCs

```{r}
df_tpiA <- df_loadings |>
    filter(gene == "tpiA") |>
    select(gene, starts_with("PC"), -PC1_abs, -PC2_abs) |>
    pivot_longer(-gene, names_to = "PC", values_to = "loading") |>
    mutate(PC = factor(PC, levels = paste0("PC", 1:8)))

p_tpiA <- ggplot(df_tpiA, aes(x = PC, y = loading)) +
    geom_bar(stat = "identity") +
    theme_classic() +
    ggtitle("tpiA loading across all PCs")
p_tpiA
```

# PCA on normalized data

```{r}
# PCA on normalized (variance-stabilized) data
# This builds on the normalization from the size factors section,
# re-creating the DESeq2 object to then apply VST and run PCA.
```

## Variance vs mean (raw)

```{r}
# There tends to be a trend in the variance, compared to mean expression values.
p <- ggplot(df_dispersion_real, aes(x=mean, y=variance)) +
    geom_point(color='grey') + 
    scale_x_log10() + scale_y_log10() + 
    theme_classic() +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)') +
    ggtitle('Spaceflight data')
p
ggsave("plots/variance_mean_real_data_raw.pdf", plot=p, width=10, height=10, units='cm')
```

## Z-score normalization

```{r}
# If we were to use z-scores
# (Often done when using PCAs)
count_data_real_long_zscore <- count_data_real_long |>
  group_by(gene) |>
  mutate(counts_z = counts/sd(counts)) |>
  ungroup()
# Plot our gene
MYGENE = "AT2G41355"
p <- count_data_real_long_zscore |> 
    filter(gene==MYGENE) |>
    ggplot(aes(x=condition, y=counts_z))+
        geom_jitter(height = 0, width = 0.1) +
        theme_classic() +
        ggtitle(MYGENE)
p
```

```{r}
# now calculate dispersions z
df_dispersion_real_zscore <- count_data_real_long_zscore |>
    group_by(gene) |>
    summarise(mean = mean(counts_z),
              sd   = sd(counts_z),
              variance = var(counts_z)) |>
    mutate(alpha = (variance-mean)/mean^2)
# now plot dispersions z
p <- ggplot(df_dispersion_real_zscore, aes(x=mean, y=variance)) +
    geom_point(color='grey') + 
    scale_x_log10() + scale_y_log10() + 
    theme_classic() +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)') +
    ggtitle('Spaceflight data, z-scores')
p
ggsave("plots/variance_mean_real_data_zscore.pdf", plot=p, width=10, height=10, units='cm')
```

```{r}
# similarly, we'd see a trend if we'd take the log
df_disperson_log <- count_data_real_long_zscore |>
    mutate(      
      log_counts   = log(counts+.1),
      log_counts_z = log(counts_z+.1)) |>
    group_by(gene) |>
    summarise(
      mean_z = mean(log_counts_z),
      sd_z   = sd(log_counts_z),
      variance_z = var(log_counts_z),
      mean = mean(log_counts), 
      sd = sd(log_counts),
      variance = var(log_counts))
# plot again
p <- ggplot(df_disperson_log, aes(x=mean, y=variance)) +
    geom_point(color='grey') + 
    scale_x_log10() + scale_y_log10() + 
    theme_classic() +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)') +
    ggtitle('Spaceflight data')
p
```

## Variance-stabilizing transformation (VST)

```{r}
# So now let's check out how dispersions look using vst
vsd <- vst(dds)
vst_counts <- assay(vsd)
df_dispersion_ddsvst <- vst_counts |>
    as.data.frame() |>
    rownames_to_column("gene") |>
    pivot_longer(-gene, names_to = "sample", values_to = "counts") |>
    group_by(gene) |>
    summarise(mean = mean(counts),
              sd   = sd(counts),
              variance = var(counts)) |>
    mutate(alpha = (variance - mean) / mean^2)
# now plot
p <- ggplot(df_dispersion_ddsvst, aes(x=mean, y=variance)) +
    geom_point(color='grey') + 
    scale_x_log10() + scale_y_log10() + 
    theme_classic() +
    xlab('Mean gene expression (all conditions)') +
    ylab('Variance in gene expression (all conditions)') +
    ggtitle('Spaceflight data, variance stabilized')
p
ggsave("plots/variance_mean_real_data_vst.pdf", plot=p, width=10, height=10, units='cm')
```

## PCA on VST data

```{r}
# now a pca plot by deseq2
pca_res <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
p <- ggplot(pca_res, aes(x=PC1, y=PC2, shape=condition,
                      color=condition)) +
    geom_point(size=3) +
    theme_classic() +
    xlab(paste0("PC1 (", round(100*pca_res$percentVar[1], 1), "%)")) +
    ylab(paste0("PC2 (", round(100*pca_res$percentVar[2], 1), "%)")) +
    ggtitle("PCA on vst data") +
    scale_color_manual(values=c("space_flight"="darkblue", "ground_control"="brown")) +
    theme(legend.position = "none")
p
ggsave("plots/pca_deseq2.pdf", plot=p, width=10, height=10, units='cm')
```

## Normalized count totals (recap)

```{r}
# Also in the overall DESeq2 analysis, dispersion-stablization is used
# (See also https://hbctraining.github.io/Intro-to-DGE/lessons/04b_DGE_DESeq2_analysis.html)

# So we have
# - Size factor correction
# - Dispersion correction

# Plot, remind ourselves that also the per-sample count normalization went well
normalized_counts_sampletotals <- colSums(normalized_counts)
p <- 
  data.frame(
    counts = unname(normalized_counts_sampletotals),
    sample = names(normalized_counts_sampletotals)
  ) |>
  mutate(sample = str_remove(sample, "Atha_WT_Col_0_sl_")) |>
  ggplot(aes(x = sample, y = counts)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("Sample") + ylab("Counts\n(normalized by size factors)")
ggsave("plots/total_counts_per_sample_normalized.pdf", plot=p, width=7, height=10, units='cm')


# Run the whole analysis
dds_2 <- DESeq(dds) # saving to dds_2 here, usually not done
```

# Statistical testing

```{r}
```